\chapter*{Abstract}
\rhead[\fancyplain{}{\bfseries Abstract}]{\fancyplain{}{\bfseries\thepage}}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries Abstract}}
\addcontentsline{toc}{chapter}{Abstract}

Reduction-based proofs are the standard technique for establishing the security of cryptographic constructions in the computational model. However, encoding such proofs within formal systems remains difficult, as they involve probabilistic computation, adversarial interaction, and explicit reasoning about computational efficiency. This thesis studies the use of programming-language techniques for formalizing cryptographic proofs in the computational model. We work within $\lamBLL$, a probabilistic lambda calculus whose graded type system characterizes probabilistic polynomial-time computation. To support reasoning about program correctness and cryptographic experiments, we introduce an assertion-annotated extension of $\lamBLL$ inspired by Hoare logic, together with a lifting procedure that embeds standard typing judgments into the enriched system. Within this framework, we formalize the security of message authentication codes constructed from pseudorandom functions by proving existential unforgeability under chosen-message attacks. The proof encodes the MAC forgery experiment, adversarial oracle access, and the standard reduction to pseudorandom function indistinguishability directly in the calculus. Overall, this work shows that $\lamBLL$ is expressive enough to represent classical cryptographic reductions while statically enforcing polynomial-time resource bounds. Moreover, by extending it with Hoare logic, it becomes possible to formalize proofs that require reasoning more complex reasoning about stateful references, such as pseudorandom-function–induced MACs, providing a possible foundation for a formal verification framework in computational cryptography.

\clearpage{\pagestyle{empty}\cleardoublepage}
