\chapter{Computational Cryptography}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries\rightmark}}
Modern cryptography is concerned with providing security guarantees for information systems, ensuring that malicious users cannot compromise their intended behaviour by corrupting data, stealing information, or disrupting system operation. Security in cryptography is therefore closely tied to formal properties defined and analysed using the language of mathematics. To study and reason about these security properties, two main lines of research exist: the computational model \cite{theory-of-computing, introduction-to-modern-cryptography} and the symbolic model, also known as the Dolev–Yao model \cite{on-the-security-of-pk-protocols}.

The computational model is tightly linked to complexity theory. Within this framework, notions such as secure encryption, unforgeable digital signatures have been placed on solid theoretical foundations. In the computational model, cryptographic primitives, security properties, protocols, and even adversaries are modeled as probabilistic polynomial-time (PPT) algorithms, reflecting adversaries that are bounded in computational resources and whose success is quantified probabilistically. To be effective, the computational model must impose reasonable restrictions when defining adversarial capabilities. The success of cryptography depends heavily on how well its security definitions capture realistic and unknown adversaries. Definitions that are too weak fail to provide meaningful guarantees, while overly strong definitions may render secure constructions impossible. For instance, in classical cryptography, a strong security result is achievable using the Vernam cipher under Shannon’s definition of perfect secrecy \cite{shannon-communication-theory-of-secrecy-systems, joyofcryptography}. However, its use in real-world settings is impractical, due to the fact that the key can be used only once (indeed, the Vernam cipher is also known as the one-time pad) and it must be at least as long as the messages that need encryption. The computational model serves as the reference framework for relatively simple protocols, but its use becomes less practical as protocol complexity increases. This is due to the fact that probabilistic reasoning becomes difficult as protocols grow larger and the number of participating parties increases.

In contrast, the symbolic model requires very few assumptions about an adversary’s computational capabilities \cite{on-the-security-of-pk-protocols}. As a result, protocol security can be proved without relying on computational assumptions. When more complex protocols are considered the computational model often reveals its limitations in terms of analysis scalability. In the symbolic model instead, once a protocol is specified, determining whether it is secure can often be automated; indeed, the symbolic model forms the basis for model checking and semi-automated theorem proving \cite{Blanchet_2022, tamarin}. The symbolic model relies on a set of predefined cryptographic primitives that are assumed to be secure and treated as idealized black boxes. Cryptographic messages as first-order terms,
together with an equational theory that represents attacker capabilities. Under these assumptions, the model is effective at identifying logical flaws in protocol designs. A well-known example is the Needham–Schroeder protocol, which was shown to be vulnerable to replay attacks \cite{timestamps-in-key-distribution-protocols}; this vulnerability can be detected vert straightforwardly using any symbolic model since it's a protocol flaw and not a cryptographic primitive vulnerability. Several formal frameworks exist for defining symbolic models, depending on the available cryptographic primitives and the underlying theory. One particularly well-studied approach is based on multi-set rewriting \cite{multiset-rewriting-and-the-complexity-of-bounded-security-protocols, rewriting-for-cryptographic-protocol-verification}. However, tools based on the symbolic model struggle to support cryptographic primitives with rich algebraic properties. For this reason, the security guarantees provided by the symbolic model are often weaker than those offered by the computational model, and a proof of security in the symbolic model does not always imply a corresponding proof in the computational model. For example, secret keys are modeled in the computational model as long bitstrings that are drawn uniformly at random while in the symbolic models they are modeled using abstract names. In symbolic models, two
distinct secret keys are represented by different names, which cannot be equal. However, in the computational model it is possible (but very unlikely) that the sampled bitstrings are equal. 

The two models have had very different outcomes with respect to the application of language-based verification techniques. The symbolic model naturally lends itself to the use of classical verification methods. In contrast, applying such techniques within the computational model is significantly more challenging, resulting in less flexibility and automation compared to the symbolic models. The connection between the two approaches (modulo additional assumptions) is established through computational soundness theorems \cite{reconciling-two-views-of-cryptography}.

Computational (or complexity theoretic) cryptography relies on the famous $\mathcal{P}$ vs $\mathcal{NP}$ probelm. If $\mathcal{P}\neq\mathcal{NP}$ and one way functions
exist, then one can derive most private-key cryptography \cite{one-way-functions-are-essential-for-complexity-base-cryptography}. If $\mathcal{P}=\mathcal{NP}$ then one way functions do not exist and all cryptographic
constructions that relies on complexity theoretic security are useless.
Cryptography is concerned with the construction of efficient schemes for which it is infeasible to violate the intended security properties. The computational model clarifies the notions of efficient and infeasible computation by formally defining the computational resources available to legitimate users and adversaries. The computations performed by honest parties are required to be efficient, while any attempt by an adversary to violate the security guarantees of the scheme should be computationally infeasible. Efficiency is commonly modeled by restricting computations to run in polynomial time in a security parameter. Adversarial computations are likewise assumed to be polynomially bounded, although the specific polynomial is not fixed in advance. Under this assumption, adversaries are restricted to the class of polynomial-time computations, and any computation requiring super-polynomial resources is considered infeasible \cite{theory-of-computing}. This framework implies a relaxation of the classical notion of perfect secrecy \cite{shannon-communication-theory-of-secrecy-systems}. Perfect secrecy requires that absolutely no information about the plaintext is leaked by the ciphertext, even to adversaries with unbounded computational power. This strong notion of security can be relaxed in order to obtain efficient protocols that are not perfectly secure in the information-theoretic sense, but are still considered secure in real-world scenarios. In the computational model, protocols may be deemed secure even if they leak a negligible amount of information to computationally bounded adversaries, provided that any such information can be exploited only with negligible probability. Security definitions that restrict adversaries by their computational power are referred to as computational security, in contrast to information-theoretic security, which makes no such assumptions \cite{introduction-to-modern-cryptography}. Of course the relaxations need very well definitions in order to capture the right amount of adversaries but still enabling efficient protocols for legitimate users. In the computational model two different approaches exits: the concrete approach and the asymptotic approach.

\section{Concrete Approach}
In the concrete approach, an encryption scheme is defined $(t, \epsilon)$-secure, if every adversary running for time at most $t$ succeeds in breaking the scheme with probability at most $\epsilon$. Of course, the notion of breaking a scheme must be precisely defined before such an analysis can be carried out. This approach differs from asymptotic computational security in that security is quantified using explicit, concrete parameters rather than asymptotic bounds. Instead of requiring security against all polynomial-time adversaries with negligible success probability, the concrete approach provides explicit guarantees relating an adversary’s running time to its maximum achievable success probability. The concrete approach is useful in practice because it yields explicit security guarantees for a cryptographic protocol, expressed in terms of the probabilities of concrete events, such as a successful key recovery. However, such guarantees are often difficult to obtain, since probability estimates typically depend on assumptions about the efficiency of the underlying hardware available to the adversary. As an illustrative example, suppose the key length is $n$, so that the key space has size $2^n$, and the key is a uniformly random bitstring of length 
$n$. A brute-force attack performed by an adversary running for $t$ CPU cycles succeeds with probability approximately $ct/2^n$, for some constant $c$ depending on the implementation. If 
$n=60$ and the adversary uses a desktop computer running at 4GHz (that is, $4\times10^9$ cycles per second), then a full brute-force search would require approximately $2^{60}/(4 \times 10^9)$ seconds which is about 9 years. A modern exascale supercomputer is capable of performing approximately $2 \times 10^{18}$ operations per second and would require it roughly $2^{60}/(2 \times 10^{18})$ seconds to break the scheme. It is on the order of half a second \cite{introduction-to-modern-cryptography}. Nonetheless a key length of $n=128$ deters even the most powerful supercomputer since $2^{128}/(2 \times 10^{18}) \approx 1,7 \times 10^{20}$ which is $5,39 \times 10^{12}$ (5,39 trillion years). This clearly show some of the major limitations of the concrete security approach, which often struggles to provide meaningful guarantees without making additional assumptions about an adversary’s hardware capabilities or the availability of highly optimized algorithms implemented on specialized hardware for specific tasks.

\section{Asymptotic Approach}
The asymptotic approach is rooted in complexity theory. All parties involved in a cryptographic protocol are modeled as algorithms that take as input a security parameter, usually denoted by 
$n$. The notion of efficiency is formalized using PPT algorithms. An efficient adversary is defined as a probabilistic algorithm whose running time is bounded by a polynomial in the security parameter; that is, there exists a polynomial $p$ such that the adversary runs in time at most $p(n)$. The notion of small \textit{probability of success} is given by the introduction of negligible functions.

\begin{definition}[Negligible Functions]
A function $f: \mathbb{N} \to \mathbb{R}$ is said to be negligible iff for every polynomial $p: \mathbb{N} \to \mathbb{N}$ there is a $N$ such that for every $n > N$ it holds that $f(n) < 1/p(n)$. 
\end{definition}

\begin{lemma}\label{lemma:closure-ngl}
   The set of all negligible functions $\mathcal{NGL}$ is closed for sum, product and product by an arbitrary polynomial.
\end{lemma}
\begin{proof}
Let $f(n)$ be a negligible function and let $q(n)$ be any polynomial that is positive for sufficiently large $n$. We must show that the function $h(n) = q(n) f(n)$ is negligible. Let $p(n)$ be an arbitrary polynomial. Define a new polynomial $p'(n) = p(n) q(n)$. Since the product of two polynomials is itself a polynomial, $p'(n)$ is a polynomial. Because $f(n)$ is negligible, there exists an integer $N$ such that
\[ f(n) < \frac{1}{p'(n)} \quad \forall n > N \]
Substituting $p'(n) = p(n) q(n)$, we obtain
\[ f(n) < \frac{1}{p(n) q(n)} \quad \forall n > N \]
Multiplying both sides by $q(n)$ yields
\[ q(n) f(n) < \frac{1}{p(n)} \quad \forall n > N \]
Because $p(n)$ was arbitrary, it follows that $h(n)$ is negligible. Closure under addition and multiplication of negligible functions follows by similar arguments.
\end{proof}

The definition of negligible functions helps formalize the extent to which an adversary may hope to break a cryptographic scheme being bounded in computational time. At first, however, the role of probability may appear misleading in a cryptographic setting. In this context, probability does not reflect uncertainty in the correctness of cryptographic algorithms such as encryption/decryption or signature/verification which are deterministic in their correctness guarantees. Rather, probability arises because all algorithms (hence parties) in a cryptographic protocol use randomized subroutines that are essential for achieving security. Indeed, without randomness, secrecy cannot be generated \cite{theory-of-computing}. For this reason, probabilistic reasoning and randomized computations are indispensable components of the computational model. 

In most cases, security proofs in the computational model are reductionist. That is, given an arbitrary but computationally bounded adversary that breaks the security of a cryptographic construction with non-negligible probability, one constructs another computationally bounded adversary that breaks an underlying hardness assumption with comparable probability. This methodology is known as provable security and enables the development of rigorous and detailed mathematical proofs. In practice, the security of a cryptographic scheme $\Pi$ is often reduced to the security of another scheme or assumption $\Pi'$. Reductionist proofs can be viewed as proofs by contraposition since a reduction establishes an implication $(P\Rightarrow Q)$, it is equivalent to proving its contrapositive $(\neg Q \Rightarrow \neg P)$. For example, if $P$ denote the statement that $\Pi$ is secure, and $Q$ denote the statement that $\Pi'$ is secure. A proof by reduction shows that if $\Pi'$ is secure, then $\Pi$ is secure. In the same way if $\Pi$ is insecure (i.e., there exists an efficient adversary that breaks $\Pi$), then $\Pi'$ is also insecure. This means that given any adversary $A$ that breaks $\Pi$ with non-negligible probability, one can construct an adversary $B$ that uses $A$ as a subroutine and breaks $\Pi'$ with non-negligible probability. Thus, an efficient attack against $\Pi$ would imply an efficient attack against $\Pi'$, contradicting the assumed security of $\Pi'$. 

As cryptography evolves, driven both by increasingly sophisticated adversaries and by the emergence of new technologies, the computational model face several challenges. In particular, it struggles to scale probabilistic reasoning and to apply formal, rigorous mathematical methods to increasingly complex protocols. These difficulties motivate the development of machine-checked frameworks that support the formal construction and automated verification of cryptographic systems. While the symbolic model naturally lends itself to model checking and semi-automated theorem proving, the computational model requires substantially more effort, since reasoning about probabilistic, resource-bounded adversaries becomes increasingly challenging as protocols grow in complexity.

\section{Perfect Secrecy}\label{sec:perfect-secrecy}
The notion of encryption scheme $\Pi$ in the computational model refers to a triple of algorithms $\Pi=(Gen, Enc, Dec)$ with a message space $\mathcal{M}$ and a key space $\mathcal{K}$ which is the set of all the possible keys that can be produced by $Gen$. The key-generation algorithm $Gen$ is a probabilistic algorithm that outputs a key $k \in \mathcal{K}$ chosen according to some distribution. The encryption algorithm $Enc$ takes as input a key $k \in \mathcal{K}$ (often compacted to $Enc_k$) and a message $m \in \mathcal{M}$, and outputs a ciphertext $c$. Finally $\mathcal{C}$ denote the set
of all possible ciphertexts that can be output by $Enc_k(m)$, for all possible
choices of $k \in \mathcal{K}$ and $m \in \mathcal{M}$. The algorithm $Dec$ is  Given the general definition of an encryption scheme, one may naturally ask whether it is secure. In the computational model, the security of an encryption scheme relies on explicit assumptions about both the computational power of the adversary and the probability with which the adversary may succeed in breaking the scheme while the scheme is still considered secure. Formalizing the notion of security in the computational model therefore constitutes the first and most fundamental set of assumptions. First of all, according to Kerckhoffs' principle, as phrased by Shannon \enquote{the enemy knows the system} \cite{shannon-communication-theory-of-secrecy-systems}, the security of an encryption scheme must not rely on the secrecy of its design or algorithms. Relying on hidden algorithms is commonly referred to as security through obscurity, a practice that is widely discouraged. The only piece of information that is always assumed to remain secret is the key $k$. This convention reflects another formulation of Kerckhoffs' principle: all of the security of a cryptographic system should be concentrated in the secrecy of the key, not in the secrecy of the algorithms themselves \cite{joyofcryptography}. If a key is leaked, generating a new one is simple and straightforward given the algorithm $Gen$, while studying and producing a new cryptographic scheme may take decades. Rather than defining security in terms of informal arguments or abstract intuition, computational cryptography formalizes security through cryptographic experiments (or games). These experiments specify the interaction between a challenger (the scheme) and an adversary and define what it means for the adversary to break a cryptographic scheme. Security is then defined by requiring that any probabilistic polynomial-time adversary succeeds in these experiments with at most negligible probability. This contrasts with the perfect-secrecy setting, where there is no notion of negligible functions: security properties rely entirely on the structural design of the scheme and its exact probabilistic behaviour, rather than on computational limitations of the adversary.

In the context of perfect secrecy, the encryption algorithm  $Enc$ may be probabilistic, while the decryption algorithm $Dec$ is deterministic, and no other computational limitations are imposed on the algorithms involved. The processes of key generation, message selection, encryption, and decryption are all viewed as parts of a joint probabilistic experiment. By introducing random variables such as $\mathbf{K}$, $\mathbf{M}$ and $\mathbf{C}$, representing respectively the key generated by $Gen$, the message chosen for encryption, and the resulting ciphertext, it becomes possible to reason formally $\mathrm{Pr}[\mathbf{K} = k]$ where $k \in \mathcal{K}$ is a key, or $\mathrm{Pr}[\mathbf{K} = k\,\,|\,\,\mathbf{M} = m]$ where $m \in \mathcal{M}$.
\begin{definition}[Perfect Secrecy]
    An encryption scheme $\Pi = (Gen, Enc, Dec)$ is said to be perfectly secret if for every message $m \in \mathcal{M}$ and every ciphertext $c \in \mathcal{C}$ such that $\mathrm{Pr}[\mathbf{C} = c] > 0$ it holds that $\mathrm{Pr}[\mathbf{M} = m\,\,|\,\, \mathbf{C} = c] = \mathrm{Pr}[\mathbf{M} = m]$
\end{definition}
The intuition behind this definition is that observing the ciphertext reveals absolutely no information about the plaintext. At the same time the distribution of the ciphertext is independent of the message being encrypted: for any two messages $m, m'\in \mathcal{M}$ the distribution of the ciphertext produced by encrypting $m$ is identical to the distribution of the ciphertext produced by encrypting $m'$. A scheme that definitely has such property is the Vernam's cipher also known as One-Time Pad (OTP) \footnote{One-time pad is often called Vernam’s cipher after Gilbert Vernam, a telegraph engineer who patented the scheme in 1919. However, an earlier description of one-time pad was recently discovered in a 1882 text by Frank Miller on telegraph encryption \cite{Bellovin01072011}.}. OTP is defined as a triple of algorithms $Gen, Enc, Dec$ reported in Figure~\ref{fig:otp}.
\begin{figure}[htpb]
    \centering
    \begin{align*}
    &\mathcal{K} = \mathcal{M} = \mathcal{C} = \{0,1\}^n &Gen(n) = k \gets \{0,1\}^n;\,\,\mathtt{return}\,\,k \\
    &Enc_k(m) = \mathtt{return}\,\, m \oplus k
    &Dec_k(c) = \mathtt{return}\,\, c \oplus k
    \end{align*}
    \caption{Definition of the OTP encryption scheme}
    \label{fig:otp}
\end{figure}

Where the notation $k \gets \{0,1\}^n$ means to sample uniformly from the set of $n$-bitstrings. This uniform choice of key is the only randomness in all of the one-time pad algorithms. The scheme is correct because $Dec_k(Enc_k(m)) = (m\oplus k)\oplus k = m \oplus (k\oplus k) = m\oplus {0}^n = m$
OTP is perfectly secure in the information-theoretic sense, but it suffers from major limitations: the key length must be greater than or equal to the length of the message, and each key can be used only once.

\newcommand{\privkeav}[0]{\mathtt{PrivK}^{eav}_{A,\Pi}}

Encryption schemes can be studied in terms of perfect secrecy, but also in terms of indistinguishability, which is an equivalent and widely used definition of secrecy for encryption schemes. To discuss the indistinguishability property of an encryption scheme, we introduce the experiment $\privkeav$.

\begin{figure}[htpb]
    \centering
    \begin{align*}
    &\privkeav(n): \\
    &\quad m_0, m_1 \gets A \\
    &\quad k \gets Gen(1^n) \\
    &\quad b \gets \{0,1\} \\
    &\quad c \gets Enc_k(m_b) \\
    &\quad b^* \gets A(c) \\
    &\quad \mathtt{return}\,\, \neg(b \oplus b^*)
    \end{align*}
    \caption{Pseudocode for the $\mathtt{PrivK}^{eav}$ experiment}
    \label{fig:privk-eav-perfect-secrecy}
\end{figure}

One may ask what the maximum probability is with which an adversary can cause the experiment to output $1$, namely
\[
\mathrm{Pr}[\mathtt{PrivK}^{Eav}_{A, \Pi} = 1]
\]
A natural question is whether it would make sense to require
\[
\mathrm{Pr}[\mathtt{PrivK}^{Eav}_{A, \Pi} = 1] = 0
\]
This requirement is clearly too strong, since an adversary $A$ can trivially succeed with probability $1/2$ by outputting a random guess. Perfect indistinguishability instead requires that no adversary can achieve a success probability strictly greater than $1/2$.

\begin{definition}[Indistinguishable Encryptions]
An encryption scheme $\Pi = (Gen, Enc, Dec)$ has indistinguishable encryptions if and only if for every adversary $A$,
\begin{align}
\mathrm{Pr}[\mathtt{PrivK}^{Eav}_{A, \Pi} = 1] = \frac{1}{2}
\end{align}
\end{definition}
It then immediately follows that:
\begin{theorem}[Perfect Secrecy Via Indistinguishability]
An encryption scheme $\Pi$ is perfectly secret if and only if it is perfectly indistinguishable.
\end{theorem}
The limitations of perfect secrecy stem from the fact that any perfectly secret encryption scheme must have a key space that is at least as large as the message space. If all keys have the same length and the message space consists of all strings of some fixed length, this implies that the key must be at least as long as the message, and this requirement is unavoidable. If two users wish to privately convey an $n$-bit message, they must first privately agree on an $n$-bit key. This also leads to the conclusion that the key length used in the one-time pad is optimal. Another important limitation is that the key can only be used once. A modified scheme, such as the so-called two-time pad, is inevitably insecure and vulnerable to attacks, most notably crib-dragging attacks. Nonetheless, the most fundamental limitation of perfect secrecy is captured by the following theorem.
\begin{theorem}
Let $\Pi = (Gen, Enc, Dec)$ be a perfectly secret encryption scheme over a message space $\mathcal{M}$ and a key space $\mathcal{K}$. Then $|\mathcal{K}| \geq |\mathcal{M}|$.
\end{theorem}
This makes perfect secrecy almost impossible in a real world setting. Just imagine the keys required to encrypt videos or very large files.

\section{Private Key Encryption}
As stated in Section~\ref{sec:perfect-secrecy}, perfect secrecy requires that absolutely no information about an encrypted message is leaked, even to an eavesdropper with unlimited computational power \cite{introduction-to-modern-cryptography}. In computational cryptography, this strong notion is weakened in favor of cryptographic schemes that are still considered secure, but only against efficient adversaries with a low probability of success. The notion of efficiency in the computational model is captured by bounding the running time of adversaries by a polynomial in a security parameter. Recall that an algorithm (or adversary) $A$ runs in polynomial time if there exists a polynomial $p$ such that, for every input $x \in {0,1}^n$, the computation of $A(x)$ terminates within at most $p(n)$ steps. The requirement of a low probability of success is formalized through the notion of negligible functions.

The computational model, which is the \emph{de facto} standard for analysing the security properties of cryptographic schemes using rigorous mathematical tools, relies on concepts from computational complexity theory. In particular, it requires that the size of inputs be well defined. For this reason, it is customary to assume, without loss of generality, that keys, messages, and ciphertexts are represented as bitstrings. Another essential requirement is that all algorithms be probabilistic. This can be viewed as each algorithm being provided, in addition to its input, with access to a uniformly distributed random tape of sufficient length, whose bits it may use as needed throughout its execution. The design choices underlying the computational model, and in particular the asymptotic approach are neither arbitrary nor accidental. One advantage of using (probabilistic) polynomial time as the measure of efficiency is that one doesn't have to precisely specify a concrete model of computation. Indeed, the extended Church-Turing thesis asserts that all reasonable models of computation are polynomially equivalent. If the analysis shows polynomial-time execution, then any reasonable implementation will also run in polynomial time. Another advantage of restricting attention to (probabilistic) polynomial-time algorithms is that they have desirable closure properties. In particular, an algorithm that makes polynomially many calls to polynomial-time subroutines, and performs only polynomially bounded additional computation, will itself run in polynomial time (see proof of Lemma \ref{lemma:closure-ngl}).

In the setting of private-key encryption, two parties share a secret key and use it to communicate confidentially. Formally, a private-key encryption scheme is a tuple of probabilistic polynomial-time algorithms $\Pi = (Gen, Enc, Dec)$, defined over a message space $\mathcal{M}$ and a key space $\mathcal{K}$, such that:
\begin{itemize}
\item The key-generation algorithm $Gen$ takes as input a string of the form $1^n$, interpreted as the security parameter, and outputs a key $k$ such that $|k| \geq n$.
\item The encryption algorithm $Enc$ may be probabilistic; that is, it may produce different ciphertexts when invoked multiple times on the same message--key pair $(m,k)$.
\item The decryption algorithm $Dec$ is deterministic, since correctness would otherwise not be guaranteed.
\end{itemize}

The scheme is assumed to be correct, meaning that for all keys $k \in \mathcal{K}$ generated by $Gen$ and all messages $m \in \mathcal{M}$ it is true that $Dec_k(Enc_k(m)) = m$. Often, the encryption algorithm $Enc_k$ is defined only for messages whose length is equal to some function $\ell(n)$ of the security parameter $n$. In this case, the encryption scheme is said to be a fixed-length encryption scheme with length parameter $\ell$.

In order to define security for such an encryption scheme let's introduce a modified notion of the experiment $\privkeav$ reported in Figure~\ref{fig:privk-eav}.

\begin{figure}[htpb]
    \centering
    \begin{align*}
    &\privkeav(n): \\
    &\quad (m_0, m_1) \gets A \\
    &\quad \mathtt{if}\,\,|m_0| \neq |m_1| \\
    &\quad\quad \mathtt{return}\,\,0 \\
    &\quad k \gets Gen(1^n) \\
    &\quad b \gets \{0,1\} \\
    &\quad c \gets Enc_k(m_b) \\
    &\quad b^* \gets A(c) \\
    &\quad \mathtt{return}\,\, \neg(b \oplus b^*)
    \end{align*}
    \caption{Pseudocode for the modified $\mathtt{PrivK}^{eav}$ experiment}
    \label{fig:privk-eav}
\end{figure}

In the experiment $\privkeav$, the adversary $A$ is given the security parameter as input and outputs a pair of messages $(m_0, m_1)$. If the two messages have different lengths, the experiment immediately returns $0$. Otherwise, a key $k$ is generated by running $Gen(1^n)$, and a uniformly random bit $b \in \{0,1\}$ is sampled. The adversary is then given the ciphertext $c = Enc_k(m_b)$ and it has to output a single bit $b^*$. The experiment returns $1$ if $b^* = b$, meaning that the adversary has successfully guessed which message was encrypted.

The definition of security in the form of a indistinguishability, for a scheme $\Pi$ with respect of $\privkeav$ is preented in Definition~\ref{def:sec-eav}.

\begin{definition}[Security Against Passive Attacks]\label{def:sec-eav}
An encryption scheme $\Pi$ is said to be secure against passive attacks or secure with respect to $\privkeav$ if and only if for every PPT adversary $A$ there exists a negligible function $\epsilon$ such that
\begin{align}
\mathrm{Pr}[\privkeav]= \frac{1}{2} + \epsilon(n)
\end{align}
\end{definition}
Once again, adversaries are restricted to probabilistic polynomial-time algorithms, and their probability of success is not required to be zero. Such a requirement would be unreasonable, since an adversary can always guess randomly and succeed with probability $1/2$. Instead, security requires that any adversary’s advantage over random guessing be negligible. The class of attacks considered with respect to the experiment $\privkeav$ consists of \emph{passive attacks}, namely attacks in which the adversary only observes information without actively interfering with the protocol execution. In particular, the adversary may have access either to a collection of ciphertexts $c_1, \dots, c_n$, or to a list of message–ciphertext pairs $(m_1, c_1), \dots, (m_n, c_n)$, where each $c_i$ is the ciphertext corresponding to $m_i$.

To construct an encryption scheme satisfying this notion of security, one must address the requirement that $|\mathcal{K}| < |\mathcal{M}|$. As seen in the case of perfect secrecy, this condition renders schemes impractical in real-world settings. There is therefore a need for algorithms that use relatively short keys while still inducing a cryptographic scheme that is secure according to Definition~\ref{def:sec-eav}. A fundamental tool for achieving this goal is the pseudorandom generator.

A pseudorandom generator is an efficient, deterministic algorithm that transforms a short, uniformly random string, called a \emph{seed}, into a longer string that appears uniform to any efficient adversary. In other words, a pseudorandom generator uses a small amount of true randomness to generate a large amount of pseudorandomness. Let $\ell : \mathbb{N} \to \mathbb{N}$ be a polynomial, called the \emph{expansion factor}, and let $G$ be a deterministic algorithm that, for every input $s \in \{0,1\}^*$, outputs a string $G(s)\in\{0,1\}^{\ell(|s|)}$. Then $G$ is a \emph{pseudorandom generator (PRG)} if and only if the following conditions hold:
\begin{itemize}
\item For every $n \in \mathbb{N}$, $\ell(n) > n$.
\item $G$ runs in probabilistic polynomial time.
\item For every PPT distinguisher $D$, there exists a negligible function $\epsilon$ such that \begin{align} |\mathrm{Pr}[D(G(s))=1]| - \mathrm{Pr}[D(r)=1] \leq \epsilon(n) \end{align}\label{eq:prng}
\end{itemize}

Here, the first probability is taken over the uniform choice of $s \in \{0,1\}^n$ and the internal randomness of $D$, while the second probability is taken over the uniform choice of $r \in \{0,1\}^{\ell(n)}$ and the internal randomness of $D$.

As stated above, a pseudorandom generator is not random in a strict information-theoretic sense. If the expansion factor $\ell$ is fixed, for example $\ell(n) = 2n$, then the set of all possible bitstrings of length $2n$ has size $2^{2n}$, the generator $G$, being deterministic, can produce at most $2^n$ distinct outputs one for each possible seed. Thus, the distribution induced by $G(s)$ has support of size at most $2^n$, which is clearly smaller than the full space $2^{2n}$. Consequently, there always exists a distinguisher $D$ that can separate the output of $G$ from the uniform distribution with high probability. However, such a distinguisher necessarily has exponential complexity. Since the definition of pseudorandomness only considers efficient (polynomial-time) distinguishers, this does not contradict Equation~\ref{eq:prng} and for all PPT distinguishers, the distinguishing advantage remains negligible.

Given the notion of a pseudorandom generator, one can construct a PRG-induced encryption scheme that is secure against passive (eavesdropping) attacks. Formally, this scheme is a triple of algorithms $\Pi^G = (Gen, Enc, Dec)$ defined as follows:
\begin{itemize}
    \item The algorithm $Gen$ on input $1^n$ outputs each string of length $n$ with same probability $1/2^n$.
    \item The algorithm $Enc$ is defined as $Enc_k(m) = m \oplus G_k$
    \item The algorithm $Dec$ is defined as $Dec_k(c) = c \oplus G_k$
\end{itemize}

The scheme $\Pi^G$ is correct because $G$ is deterministic. Indeed, for every key $k$ and message $m$ it is true that $Dec_k(Enc_k(m)) = Dec_k(m \oplus G_k) = ((G_k \oplus m) \oplus G_k)  = m$. The scheme $\Pi^G$ is defined for messages of fixed length $\ell(n)$, where $\ell$ is the expansion factor of the pseudorandom generator $G$.

\newcommand{\privkmult}[0]{\mathtt{PrivK}^{mult}_{A,\Pi}}

In the context of the experiment $\privkeav$, the considered attacks are \emph{passive}. This means that the adversary cannot actively interfere with the execution of the protocol, but can only observe its surroundings. In particular, the communicating parties transmit a single ciphertext, which may be observed by an eavesdropper. It would be desirable, however, for the communicating parties to be able to exchange multiple ciphertexts generated using the same secret key even in the presence of an eavesdropper who can observe all of them. To model such scenarios an encryption scheme is required to remains secure under the encryption of multiple messages, and has to behave accordingly with respect of a new security experiment, shown in Figure~\ref{fig:privk-mult}.

\begin{figure}[htpb]
    \centering
    \begin{align*}
    &\privkmult(n): \\
    &\quad (\vec{M_0}, \vec{M_1}) \gets A \\
    &\quad \mathtt{if}\,\,\exists i. |m_{0,i}| \neq |m_{1,i}| \\
    &\quad\quad \mathtt{return}\,\,0 \\
    &\quad k \gets Gen(1^n) \\
    &\quad b \gets \{0,1\} \\
    &\quad \vec{C} \gets Enc_k(\vec{M}_b) \\
    &\quad b^* \gets A(c) \\
    &\quad \mathtt{return}\,\, \neg(b \oplus b^*)
    \end{align*}
    \caption{Pseudocode for the $\mathtt{PrivK}^{mult}$ experiment}
    \label{fig:privk-mult}
\end{figure}
In the experiment $\privkmult$, the adversary is allowed to output two vectors of messages of length $t$, namely $\vec{M_0} = (m_{0,0}, \dots , m_{0,t-1})$ and $\vec{M_1} = (m_{1,0}, \dots , m_{1,t-1})$. A random bit $b$ is chosen, and the challenger encrypts the entire vector $\vec{M}_b$ under the same key $k$, producing a ciphertext vector $\vec{C} = (c_0, \dots, c_{t-1})$ which is then given to the adversary.
\begin{definition}[Security Against Multiple Encryptions]
A scheme $\Pi$ is said to be secure against multiple encryptions if for every PPT adversary $A$ there exists a negligible function $\epsilon$ such that
\begin{align}
\Pr[\privkmult = 1] \leq \tfrac{1}{2} + \epsilon(n).
\end{align}
\end{definition}

It immediately follows that:

\begin{lemma}
The scheme $\Pi^G$ is not secure with respect to $\privkmult$, even if $G$ is a pseudorandom generator.
\end{lemma}

What at first appears to be a strong limitation is in fact fundamental.

\begin{theorem}
If the encryption algorithm $Enc$ is stateless and deterministic, then the scheme $\Pi = (Gen, Enc, Dec)$ cannot be secure with respect to $\privkmult$.
\end{theorem}

\newcommand{\privkcpa}[0]{\mathtt{PrivK}^{cpa}_{A,\Pi}}

Another important class of attacks consists of \emph{active} attacks, in which adversaries are allowed to play an active role. In this setting, the adversary may obtain encryptions of messages of its choice by interacting with an encryption oracle $Enc_k(\cdot)$ (chosen-plaintext attacks), or may even have access to a decryption oracle $Dec_k(\cdot)$ (chosen-ciphertext attacks). In both cases, the secret key $k$ is unknown to the adversary. Chosen-plaintext attacks (CPA) are particularly useful for introducing a stronger and more realistic security notion. In this scenario, the adversary $A$ is allowed to interact with the encryption oracle and query it on messages of its choice, for a polynomial number of queries in the security parameter $n$. This interaction is formalized by the experiment shown in Figure~\ref{fig:privk-cpa}.
\begin{figure}[htpb]
    \centering
    \begin{align*}
    &\privkcpa(n): \\
    &\quad k \gets \{0,1\}^n \\
    &\quad (m_0, m_1) \gets A^{Enc_k(\cdot)}(1^n) \\
    &\quad b \gets \{0,1\} \\
    &\quad c \gets Enc_k(m_b) \\
    &\quad g \gets A(c, Enc_k(\cdot)) \\
    &\quad \mathtt{return}\,\,(b=g)
    \end{align*}
    \caption{Pseudocode for $\mathtt{PrivK}^{cpa}$ experiment}
    \label{fig:privk-cpa}
\end{figure}
In the experiment $\privkcpa$, the adversary first obtains access to the encryption oracle and outputs two messages $m_0$ and $m_1$ of equal length. A random bit $b$ is then sampled, and the challenge ciphertext $c = Enc_k(m_b)$ is returned to the adversary. The adversary continues to have access to the encryption oracle and must output a guess $g$ for the value of $b$. When the algorithm $Enc$ is randomized, the
oracle uses fresh randomness each time it answers a query. 

Security against chosen-plaintext attacks requires that no efficient adversary can guess the value of $b$ with probability significantly greater than $1/2$. This is captured by the following definition.
\begin{definition}[CPA Security]
An encryption scheme $\Pi$ is said to be secure against chosen-plaintext attacks if and only if for every PPT adversary $A$ there exists a negligible function $\epsilon$ such that
\begin{align}
\Pr[\privkcpa(n) = 1] \leq \frac{1}{2} + \epsilon(n)
\end{align}
where the probability is taken over the randomness used by A, as well as the randomness used in the experiment.
\end{definition}

An immediate consequence of the definition of security with respect to $\privkcpa$ is that the encryption algorithm must be probabilistic. This observation reinforces the need to rely on cryptographic primitives that introduce randomness into the encryption.
\begin{lemma}
Any encryption scheme $\Pi = (Gen, Enc, Dec)$ that is secure with respect to $\mathtt{PrivK}^{cpa}$ must have a probabilistic encryption algorithm $Enc$.
\end{lemma}
This requirement also provides a solution to the problem of security under multiple encryptions, which was shown to be unachievable when relying only on pseudorandom generators. Indeed, probabilistic encryption achieve security even when the same key is used to encrypt multiple messages, as captured by the following theorem.
\begin{theorem}
Every encryption scheme that is CPA-secure is also secure in the presence of multiple encryptions.
\end{theorem}

To construct a CPA-secure encryption scheme, it is necessary to introduce the notion of a \emph{pseudorandom function} (PRF). Informally, a pseudorandom function captures the idea of pseudorandomness for a distribution over functions. Such a distribution is induced by considering \emph{keyed functions}, defined as follows. A keyed function is a two-input function $F: \{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ where the first input is the key and is denoted by $k$. Typically a key $k$ is chosen once and fixed. One then considers the single-input function $F:\{0,1\}^* \to \{0,1\}^*$ denoted by $F_k$. By choosing a key $k \in \{0,1\}^n$ uniformly at random, the keyed function $F$ induces a distribution over single-input functions $F_k$, so a PRF is not a single function, but a family of functions indexed by a key, for each key fixed there is a function. The intuition is that $F$ is pseudorandom if the distribution is computationally indistinguishable from the uniform distribution over all functions with the same domain and range. Meaning that no efficient adversary should be able to tell whether it is interacting with $F_k$ for a uniformly chosen key $k$, or with a truly random function $f$. 

A very important consideration concerns the size of the set of truly random functions from $\{0,1\}^n$ to $\{0,1\}^n$. The number of all such functions is $2^{n 2^n}$ since for each of the $2^n$ possible inputs one may independently choose any of the $2^n$ possible outputs. By contrast, when fixing a key $k \in \{0,1\}^n$ for a keyed function $F$, the resulting function $F_k$ is just a single deterministic function. As $k$ ranges over all $n$-bit strings, the family ${F_k}$ contains at most $2^n$ distinct functions. This induces a distribution over functions supported on at most 
$2^n$ functions if all keys define distinct functions, then each occurs with probability $1/2^n$. Requiring that no efficient adversary be able to distinguish a truly random function from a pseudorandom function amounts to an extremely strong requirement. It asks the adversary to be unable to distinguish between a uniform distribution over an astronomically large set of size $2^{n 2^n}$ where each function has probability $1/2^{n 2^n}$ and a distribution supported on \enquote{only} $2^n$ functions, each occurring with probability $1/2^n$. These two distributions are vastly different. Nevertheless, the defining property of a pseudorandom function is that this difference cannot be exploited by any PPT adversary.

To formally define pseudorandom function let's introduce also the following notions
\begin{itemize}
    \item A binary partial function is a partial function from $\{0,1\}^n$ to $\{0,1\}^n$
    \item A binary partial function $F$ is length-preserving iif $F(k,x)$ is defined iif $|k|= |x|$ and in that case $|F(k,x)| = |x|$.
    \item A binary partial function efficient if there exists a PPT algorithm that computes it.
\end{itemize}
\begin{definition}
Let $F: \{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ be an efficient, length-preserving, keyed function. $F$ is a pseudorandom function if for all PPT distinguishers $D$, there is a negligible function $\epsilon$ such that:
\begin{align}
\mathrm{Pr}[D^{F_k(\cdot)}(1^n) = 1] - \mathrm{Pr}[D^{f(\cdot)}(1^n) = 1] \leq \epsilon(n)
\end{align}
where $k$ is  uniformly chosen among all strings of length $n$ and $f(\cdot)$ is uniformly chosen among all functions from $\{0,1\}^n$ to $\{0,1\}^n$.
\end{definition}

Given a pseudorandom function, one can construct a private-key encryption scheme that is secure with respect to the experiment $\mathtt{PrivK}^{cpa}$. The resulting scheme, denoted by ($\Pi^F = (Gen, Enc, Dec)$), is defined as follows.
\begin{itemize}
    \item The algorithm $Gen$ on input $1^n$ outputs each string of length $n$ with same probability $1/2^n$.
    \item Given a pseudorandom function $F$, the algorithm $Enc_k(m)$ is defined as the (binary encoding of the) pair $\langle r, F_k(r) \oplus m\rangle$, where $r$ is a random string $|k|$ bits long.
    \item $Dec_k(c)$ returns $F_k(r) \oplus s$ anytime $c$ is the (binary encoding of the) pair $\langle r, s\rangle$.
\end{itemize}

\begin{theorem}
If $F$ is a PRF, then $\Pi^F$ is secure against CPA attacks.
\end{theorem}

\section{Message Authentication Codes}
Message Authentication Codes (MACs) address the problem of message authentication over insecure channels. In this setting, messages are not required to be secret, but it must be possible to authenticate them, namely, each party should be able to verify that a message it receives was indeed sent by the party claiming to be the sender. In other words, MACs provide integrity and authenticity rather than confidentiality. 

Schemes based solely on pseudorandom generators are insufficient to solve this problem. For instance, consider an encryption scheme $\Pi^G$ based on a pseudorandom generator $G$, where encryption is defined as $Enc_k(m) = G_k \oplus m$ and decryption as $Dec_k(m) = G_k \oplus c$. An adversary who observes a ciphertext can easily construct a valid ciphertext for a related message, hence impersonating the legitimate sender. Even more directly, flipping any bit in the ciphertext $c = G_k \oplus m$ results in the same bit being flipped in the plaintext after decryption. Thus, given a ciphertext $c$ that encrypts a message $m$, it is possible to generate a modified ciphertext $c'$ such that $m' = Dec_k(c')$ is equal to $m$ except for one or more flipped bits. This illustrates that pseudorandomness alone does not guarantee message integrity. Authentication requires dedicated mechanisms beyond encryption.

\newcommand{\vrfy}[0]{V\!r\!f\!y}

Two users who wish to communicate in an authenticated manner begin by generating and securely sharing a secret key $k$ prior to communication. When one party wants to send a message $m$ to the other, she computes a MAC tag (or simply a tag) $t$ based on the message and the shared key, and sends the pair $(m, t)$ to the receiver. The tag is computed using a tag-generation algorithm, denoted by $Mac$. That is, the sender computes the tag $t$ as $Mac_k(m)$ and transmits $(m, t)$. Upon receiving $(m, t)$, the receiver checks whether $t$ is a valid tag for the message $m$ with respect to the key $k$. Verification is performed by a deterministic algorithm$\vrfy$, which takes as input the key, the message, and the tag, and outputs a bit indicating acceptance or rejection of the tag on the message.

Formally, a MAC is a triple $\Pi = (Gen, Mac, \vrfy)$ of PPT algorithms such that:
\begin{itemize}
\item $Gen$ takes as input a unary representation of the security parameter $1^n$ and outputs a secret key $k$, where typically $|k| \geq n$;
\item $Mac$ takes as input a key $k$ and a message $m$, and outputs a tag $t$;
\item $\vrfy$ takes as input a key $k$, a message $m$, and a tag $t$, and outputs a bit $b$, with $b = 1$ meaning valid and $b = 0$ meaning invalid.
\end{itemize}

A MAC is said to be correct if for all keys $k$ generated by $Gen$ and all messages $m$, it holds that

\[
\vrfy_k(m,Mac_k(m))=1
\]

Correctness ensures that honestly generated tags are always accepted by the verification algorithm.

\newcommand{\macforge}[0]{\mathtt{MacForge}_{A,\Pi}}

The security goal for message authentication codes (MACs) can be stated informally as follows: no efficient adversary should be able to forge a valid tag for any message that was not previously sent and authenticated by one of the legitimate communicating parties. Equivalently, even after interacting with the system, the adversary should not be able to produce a new message–tag pair that will be accepted as valid. More precisely, the adversary is given oracle access to $Mac_k(\cdot)$ and may query the oracle to obtain valid pairs $(m,t)$, where $m$ is a message and $t$ is its corresponding tag. Any pair obtained directly from the oracle is, by definition, not considered a forgery. In practice, this models an adversary who passively observes the communication between honest parties and is able to see all transmitted messages together with their associated MAC tags. The security requirement is that, despite this access, any PPT adversary cannot generate a valid tag for any new message that was not previously authenticated. This can be formalized through an experiment called $\mathtt{MacForge}$ which is described in Figure~\ref{fig:macforge}.

\begin{figure}
    \centering
    \begin{align*}
    &\macforge(n): \\
    &\quad k \gets Gen(1^n) \\
    &\quad (m,t) \gets A^{Mac_k(\cdot)}(1^n) \\
    &\quad \mathbb{Q} \gets \{m \ | \ A \ \text{queries} \ Mac_k(m) \} \\
    &\quad \mathtt{return}\,\,(m \notin \mathbb{Q} \land \vrfy_k(m,t) = 1)
    \end{align*}
    \caption{Pseudocode for $\mathtt{MacForge}$ experiment}
    \label{fig:macforge}
\end{figure}

Given the experiment $\mathtt{MacForge}$ the notion of security for a scheme MAC $\Pi$ is captured by Definition~\ref{def:mac-security}.

\begin{definition}[MAC security]\label{def:mac-security}
A MAC $\Pi$ is secure if and only if for every PPT adversary $A$ there exists a negligible function $\epsilon$ such that 
\begin{align}
\mathrm{Pr}[\macforge = 1] \leq \epsilon(n)
\end{align}
\end{definition}
This definition of MAC security is simultaneously strong and weak. On the one hand, it is strong because it considers the scheme broken even if the adversary forges a valid tag for a message that is meaningless. Any successful forgery, regardless of the message content, translates to a violation of security. On the other hand, the definition is weak in that it does not protect against replay attacks. An adversary who records a previously transmitted and authenticated pair $(m,t)$ can resend the same pair at a later time, and it will still be accepted as valid by the receiver. Since replayed messages are not considered new forgeries under the standard definition, preventing such attacks requires additional mechanisms, such as nonces, timestamps, or sequence numbers, which go beyond basic MAC security.

A well-known result in cryptography states that pseudorandom functions can be used to construct secure message authentication codes. Given a pseudorandom function $F$, the construction $\mathrm{MAC}^F$ yields a secure MAC. Formally, let $\Pi^F = (Gen, Mac, Vrfy)$ be the triple of algorithms defined as follows:
\begin{itemize}
\item The key-generation algorithm $Gen$ takes as input the security parameter $1^n$ and outputs a uniformly random bitstring $k \in {0,1}^n$, each with probability $1/2^n$;
\item The tag generation algorithm is defined as $Mac_k(m) = F_k(m)$;
\item The verification algorithm is defined as $\vrfy_k(m,t) = (F_k(m) \stackrel{?}{=} t)$.
\end{itemize}

\begin{theorem}[Security of $\mathrm{MAC}^F$]
If $F$ is a pseudorandom function, then the MAC scheme $\Pi^F$ is a secure fixed-length message authentication code for messages of length $n$.
\begin{proof}
And then it's proven
\end{proof}
\end{theorem}