\chapter{Introduction}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries\rightmark}}

Cryptography has historically been regarded as a form of art, grounded in humanity's need to communicate while concealing information from unauthorized parties. For much of its history, research in cryptography was conducted under conditions of secrecy, primarily because cryptographic techniques conferred substantial military and political advantages. As a result, developments in cryptography often emerged as responses to events or activities that were either secret at the time or became public only much later \cite{the-code-book}.

The history of cryptography is a continuous battle between codemakers (cipher makers) and codebreakers (cipher breakers). The creation of strong ciphers, as well as the decryption of existing ones, relied heavily on creativity, inspiration, and occasional strokes of genius \cite{introduction-to-modern-cryptography}. In the absence of a rigorous theoretical foundation, the discipline lacked formal security definitions and systematic criteria to assess whether a given cryptographic construction was secure or vulnerable. These characteristics reinforced the perception of cryptography as an art rather than a formal science for many centuries. However, between the 1970s and 1980s, the spread of the internet and the availability of affordable digital hardware brought cryptography into everyday life. At the same time, theoretical advances \cite{new-directions-in-cryptography, a-method-for-obtaining-digital-signatures, probabilistic-encryption} in information theory and computer science laid the foundations for provably secure cryptosystems, transforming cryptography into a scientific discipline. Cryptography may thus be seen as a science derived from art, an appealing definition, but one that does not fully capture what cryptography has become today. Modern cryptography is a field at the intersection of mathematics and computer science, concerned with techniques for ensuring data integrity, exchanging secret keys, authenticating users, enabling electronic auctions and elections, performing secure distributed computations, implementing digital cash, establishing server identities, and much more \cite{introduction-to-modern-cryptography}. The scope of modern cryptography has expanded significantly, whereas classical cryptography focused almost exclusively on secure communication \cite{theory-of-computing}. 

While cryptography yields concrete mechanisms for addressing these challenges, such mechanisms are more appropriately regarded as applications of the discipline rather than the discipline itself. At its foundation, cryptography aims to resolve precisely formulated problems by abstracting away many of the complexities of real-world environments. This abstraction is essential, as cryptographic guarantees can only be established relative to explicitly specified adversarial models and clear assumptions. Although this may initially seem restrictive, the reliance on rigorous security definitions, formally stated assumptions, and mathematically sound proofs enables cryptography to provide robust and dependable security guarantees \cite{joyofcryptography}. This constitutes one of the main accomplishments of theoretical computer science \cite{theory-of-computing}.

At the foundation of the science of cryptography lies the notion of security. One may ask under which conditions a given cryptographic primitive or protocol can be regarded as secure, and what the term \textit{secure} precisely denotes in this context. The definition of security depends on the underlying formal framework, with two principal paradigms being the computational model and the symbolic model.

In the computational model, adversaries are modeled as probabilistic polynomial-time Turing machines. Cryptographic operations are represented as functions on binary strings, and security properties are expressed in terms of the probability that an adversary succeeds together with the computational complexity of the attack. Security is achieved when this probability is negligible. By contrast, the symbolic model abstracts away from complexity and probability, and instead relies on classical verification techniques. This makes it particularly suitable for detecting design flaws in cryptographic protocols, and it has led to most of the automated verification tools available today. However, security guarantees obtained in the symbolic model are generally weaker than those provided by the computational model.

While the symbolic model has been widely studied and integrated into formal verification frameworks, the computational model has not reached the same level of development in this respect. A promising direction for enabling the use of classical verification techniques in the computational setting is to apply theories of program equivalence to programming languages designed to capture the central complexity-theoretic notion of probabilistic polynomial-time computation. Within such languages, the notion of computational indistinguishability \cite{Goldreich_2001} can be reinterpreted as a form of observational equivalence. This reinterpretation opens the way to studying computational indistinguishability using established tools from programming language theory, such as logical relations.

The objective of this work is to employ a specific language, namely $\lamBLL$~\cite{lago2024computationalindistinguishabilitylogicalrelations}, whose main features are summarized in Chapter~\ref{chapter:lambll}, in order to formulate and formally prove an important cryptographic result: the security of fixed-length Message Authentication Codes (MACs), in particular existential unforgeability under chosen-message attacks. To this end, we introduce a Hoare logic that extends the type system of $\lamBLL$ with decorated terms, annotated with propositions that carry stateful meaning with respect to the execution of $\lamBLL$ programs. By reasoning within this enriched typing system, we also define a lifting procedure that maps typing judgments from the standard $\lamBLL$ system to the decorated system with Hoare assertions. Furthermore, we extend the set of function symbols and equations of $\lamBLL$ to reason about equivalences within the assertion-decorated system. In doing so, we provide a fully formal proof of MAC security within the $\lamBLL$ framework.

To place this work in context, Chapter~\ref{chapter:computational-cryptography} presents the necessary background on computational cryptography, in particular on information-theoretic security, pseudorandom generators, pseudorandom functions, and MACs. Chapter~\ref{chapter:formal-tools} briefly surveys automated formal verification tools in the computational model, together with their underlying logics, in order to offer a perspective on the current state of the art in formal verification in the computational setting. In Chapter~\ref{chapter:lambda-calculus} we review the lambda calculus and its main properties, and in Section~\ref{sec:typed-lambda calculus} we recall the simply typed lambda calculus, which provides the foundation for understanding and reasoning about $\lamBLL$. Chapter~\ref{chapter:lambll} revisits the main features of $\lamBLL$ and presents its extension via Hoare logic and a new set of typing rules, together with a technique for lifting $\lamBLL$-terms into the Hoare-logic extension, yielding assertion-typed terms. Finally, in Chapter~\ref{chapter:proof} we formalize, within the $\lamBLL$ framework, the proof of MAC security introduced in Section~\ref{subsec:mac-security}.