\chapter{Proving PRF-Induced MAC Security Equationally}\label{chapter:proof}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries\rightmark}}
This chapter demonstrates that existential unforgeability under chosen message attacks for pseudorandom function based MACs can be proved within the $\lamBLL$ framework. We formalize the standard reduction from MAC forgery to pseudorandom function indistinguishability by encoding the $\macexp$ experiment, the adversary, and the MAC oracle as $\lamBLL$ terms. By reconstructing the proof of Theorem~\ref{proof:handwavy} using the equational reasoning principles of the calculus, and by exploiting its extension with Hoare logic and assertion-typed terms, we demonstrate that $\lamBLL$ is sufficiently expressive to internalize classical security proofs while simultaneously enforcing strict polynomial-time constraints via its graded type system.

We begin by recalling the MAC forgery experiment, $\macexp$, as defined in Figure~\ref{fig:macforge-recall}. To bridge the gap between pseudocode and formalization, we specify the types for each component as they appear in $\lamBLL$. The concrete implementation of the terms modeling the key generation
algorithm $\mathit{Gen}$ and the MAC oracle $\mathit{Mac}_k(\cdot)$ depends on the specific MAC scheme under consideration. For this reason, these terms are left abstract and will be instantiated only after fixing a scheme $\Pi$. In the following paragraphs, we make this instantiation explicit for the schemes $\Pi^F$ and $\widetilde{\Pi}$ which are used in the proof of Theorem~\ref{theorem:mac-security}. 

\begin{figure}[htpb]
    \centering
    \begin{align*}
    &\macforge(n): \\
    &\quad k \gets Gen(1^n) \\
    &\quad (m,t) \gets A^{Mac_k(\cdot)}(1^n) \\
    &\quad \mathbb{Q} \gets \{m \ | \ A \ \text{queries} \ Mac_k(m) \} \\
    &\quad \mathtt{return}\,\,(m \notin \mathbb{Q} \land \vrfy_k(m,t) = 1)
    \end{align*}
    \caption{Pseudocode for $\mathtt{MacForge}$ experiment.}
    \label{fig:macforge-recall}
\end{figure}

We now describe how the components of the experiment are typed in $\lamBLL$. The key generation algorithm $Gen$ is a probabilistic computation that yields a secret key of polynomial length. The adversary, $Adv$, is a higher-order computation having access to an oracle. The oracle itself is controlled by the graded modality $\bang_q$, which enforces a polynomial bound on the number of queries.
The key generation algorithm $Gen$ is modeled as a probabilistic computation that,
given the security parameter, produces a secret key whose length is polynomial in $i$. The adversary is modeled as a higher-order computation that receives oracle access to the MAC function and outputs a candidate forgery. The MAC oracle itself is represented as a term whose use is controlled by a graded modality, enforcing a polynomial bound on the number of queries.

\begin{align*}
&\vdash_v^\Theta Gen: \UU \multimap \SS[i] \\
&\vdash_v^\Theta Mac: (\SS[i] \otimes \SS[i]) \multimap \SS[i] \\
&\vdash_c^\Theta Oracle: \SS[i] \multimap \bang_{q}\bigl(\SS[i] \multimap \SS[i]\bigr)\\
&\vdash_c^\varnothing Adv: \bang_{q}\bigl(\SS[i] \multimap \SS[i]\bigr) \multimap (\SS[i] \otimes \SS[i])
% &\vdash_v^\Theta \vrfy: (\SS[i] \otimes \SS[i] \otimes \SS[i]) \multimap \BB \\
% \vdash_c^\Theta &: 
% \bang_{q}\bigl(\SS[i] \multimap \SS[i]\bigr) \multimap \bigl((\SS[i] \otimes \SS[i]) \otimes \SS[r]\bigr) && \tag{Type for $Adv$}
\end{align*}

The typing judgments for the main components of the experiment are given above. Here, $i$ is a polynomial variable and $q$ denotes a polynomial in the security parameter. The oracle type $\SS[i] \multimap \bang_{q}(\SS[i] \multimap \SS[i])$ models a MAC oracle which,
once instantiated with a key, provides a function mapping messages of length $i$ to tags of the same length. The graded modality $\bang_{q}$ enforces a polynomial upper bound on the number of oracle queries that the adversary may perform. In addition to answering MAC queries, the oracle maintains a record of all messages queried by the adversary. This information encodes the set
\[
\mathbb{Q} \gets \{\, m \mid A \text{ queries } Mac_k(m) \,\}
\]
as presented in Figure~\ref{fig:macforge-recall}. To model this behavior in $\lamBLL$, we rely on the notion of a \emph{ledger}, which allows mutable storage with a statically bounded size. The reference $Q$ has type $Q : \SS[q \times (i+1)]$ and can be viewed as a table with $q$ rows, where each row consists of a message of length $i$ together with a final bit indicating whether the row is active. The size of the ledger reflects the fact that the oracle can be queried at most $q$ times. At each oracle invocation on a message $m$, the oracle updates the ledger stored in $Q$ as follows: it scans the ledger for an active row containing $m$; if none exists, it writes $m$ into the first inactive row and marks that row as active. In this way, the reference $Q$ always encodes exactly the set $\mathbb{Q}$ of messages previously queried by the adversary. To inspect the adversary's query history, we introduced in Section~\ref{sec:useful-functional-symbols-and-equations} some useful functions symbols. In particular in Table~\ref{table:new-function-symbols} we defined the function $\mathtt{contains}$, whose purpose is to determine whether a given message occurs in the encoding of the set $\mathbb{Q}$ stored in the reference context. More precisely, $\mathtt{contains}$ checks whether a bitstring of type $\SS[i]$ appears in a \textit{query ledger} of type $\SS[q\times(i+1)]$. It takes as input the stored query history together with a message and returns a boolean value.

The adversary $Adv$ takes as input the closure obtained by currying and encapsulating the key within the $Oracle$ term. Based on its interaction with the oracle, it produces a candidate forgery $(m,t)$ of type $(\SS[i] \otimes \SS[i])$. 

The verification function $\vrfy$ is not introduced since the validity of a proposed tag can be checked just by recomputing the MAC on the message and comparing it with the supplied tag.

It is worth noting that the type derivations of the terms introduced so far are parameterized by a reference context $\Theta$, which varies depending on the cryptographic experiment considered. In the model for the scheme $\Pi_F$, the context $\Theta$ contains the reference $Q$ used to record oracle queries so the types of the terms in Figure~\ref{fig:macforgeF-subterms} must be adapted by defining a reference context $\Xi=\{\,Q: \SS[q\times(i+1)]\,\}$ and setting $\Theta$ accordingly. In contrast, in the model for the scheme $\widetilde{\Pi}$, the context $\Theta$ is instantiated with a reference context containing also the ledger used to model the random function $f$, namely $\Upsilon=\{\,rand : \SS[(q+1)\times(2i+1)], Q: \SS[(q\times(i+1)]\,\}$.

We define the $\lamBLL$ term corresponding to the forgery experiment, shown in Figure~\ref{fig:macforge-term}. The experiment begins by generating a fresh key $k$ and constructing the corresponding oracle $o$. The adversary $Adv$ is then given access to $o$ and returns an attempted forgery $(m,t)$. The experiment first checks  whether the message $m$ appears in the query set $\mathbb{Q}$ by reading the reference $Q$ containing the query ledger and applying the predicate $\mathtt{contains}$. If verification succeeds the experiment immediately returns $\mathtt{false}$, otherwise if it succeeds, the experiment tests if the tag provided by the adversary is equal to the recomputed tag on the same message obtained through the oracle. The experiment returns $\true$ if $m$ was not previously queried, and $\false$ otherwise. If verification fails, the experiment immediately returns $\false$.

\begin{figure}[ht]
    \centering
    \begin{align*}
    &MacForge \triangleq \\
    &\quad \letin{k = Gen \star}{} \\
    &\quad \letin{o = Oracle\, k}{} \\
    &\quad \letin{\lis{m,tag} = Adv\, o}{} \\
    &\quad \letin{queries = \get Q}{} \\
    &\quad \letin{c=\mathtt{contains}(queries, m)}{} \\
    &\quad \mathtt{if}\, c\, \mathtt{then}\, \return \false \\
    &\quad \mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag)}
    \end{align*}
    \caption{$\lamBLL$ term for the $\mathtt{MacForge}$ experiment.}
    \label{fig:macforge-term}
\end{figure}
In order to underline the similarity with the pseudocode given in Figure~\ref{fig:macforge-recall}, we apply the syntactic sugar introduced in Equation~\ref{eq:sugar-return} to the first line of the term shown in Figure~\ref{fig:macforge-term}, observing that $Gen$ is a value.

We have to define a $\lamBLL$ term for the experiment using the encryption scheme $\Pi_F$ which in Proof~\ref{proof:handwavy} corresponds to the experiment $\mathtt{MacForge}_{A, \Pi_F}$ and one for the experiment using the encryption scheme $\widetilde{\Pi}$ which corresponds to $\mathtt{MacForge}_{A,\widetilde{\Pi}} $. In both the cryptographic experiment and the distinguisher, we need to construct two instances of the $\lamBLL$ term $\mathit{MacForge}$ (Figure~\ref{fig:macforge-term}) that models these components, namely one that interacts with the pseudorandom function $F$ and one that interacts with a truly random function $f$. 

\begin{figure}[ht]
    \begin{align*}
        \begin{aligned}
            &Gen^F \triangleq \\
            &\quad\lambda u. \\
            &\quad\letin{q=\mathtt{initQ}}{} \\
            &\quad \mathtt{set}\,Q\,q; \\
            &\quad\letin{k=\random}{} \\
            &\quad\return k
        \end{aligned}
        \qquad\qquad\qquad
        \begin{aligned}
            &\mathit{Mac}^F \triangleq \\
            &\quad\lambda x. \\
            &\quad\letin{\lis{k,m}=x}{} \\
            &\quad\return F\lis{k,m}
        \end{aligned}
    \end{align*}
    
    \begin{align*}
        \begin{aligned}
            &Oracle^F \triangleq \\
            &\quad \return(\lambda k.\return \bang(\return \lambda m. \\
            &\quad\quad \letin{q=\get Q}{} \\
            &\quad\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad\quad \mathtt{if}\ \mathtt{not}(t)\, \mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad\quad \mathit{Mac}^F\lis{k,m}))
        \end{aligned}
    \end{align*}
    \caption{Models for subterms of $MacForge^F$.}
    \label{fig:macforgeF-subterms}
\end{figure}

The model for the experiment $\mathtt{MacForge}_{A,\Pi_F}$, which we denote by $\mathit{MacForge}^F$, is obtained by instantiating the generic term $\mathit{MacForge}$ with the components $Gen^F$, $Mac^F$, and $Oracle^F$ defined in Figure~\ref{fig:macforgeF-subterms}. Unlike the pseudocode presented in Figure~\ref{fig:macforge-recall}.

The key generation algorithm of $\Pi_F$ is modeled by the term $Gen^F$, which samples a fresh key using the function symbol $\random$, whose type is $\SS[i]$. The term $\mathit{Gen}^F$ is additionally responsible for the initialization of query ledger which is instantiated by setting the value of the reference $Q$ using the function symbol $\mathtt{initQ}$ which generates a string of length $q\times(i+1)$ filled entirely with zeros.  

The pseudorandom function used by the scheme is abstracted as a generic value $F$, typed as $\vdash_v^\Theta F: (\SS[i] \otimes \SS[i]) \multimap S[i]$ reflecting the fact that the construction is parameterized by an arbitrary pseudorandom function which can be instantiated in many ways. 

The tag generation algorithm of $\Pi_F$ is modeled by the term $\mathit{Mac}^F$. Given a key $k$ and a message $m$ it computes the tag $t$ by evaluating the pseudorandom function $F$ on the pair $(k,m)$, finally returning a tag. 

The oracle $Oracle^F$ encapsulates the secret key and provides access to the MAC function through the graded modality $\bang$. In addition to answering MAC queries, the oracle maintains a record of all messages queried by the adversary. This record is stored in the reference $Q$ thus introducing a non-empty reference context. At each oracle invocation on a message $m$, the current contents of $Q$ are retrieved, updated to include $m$, and written back to the reference before computing and returning the corresponding MAC value. In order to retrieve and update the query ledger, we rely on several function symbols such as $\mathtt{initQ}$, $\mathtt{modifyQ}$ and $\mathtt{contains}$, which are reported in Table~\ref{table:new-function-symbols}. To support this behavior, the typing of the terms in Figure~\ref{fig:macforgeF-subterms} is carried out in a reference context $ \Xi = \{\, Q : \SS[q\times(i+1)] \,\}$ which represents the ledger used to store the adversary's oracle queries. The context $\Theta$ in the corresponding type derivations is instantiated with $\Xi$. In the term $Oracle^F$ in Figure~\ref{fig:macforgeF-subterms} we use the syntactic sugar of Equation~\ref{eq:sugar-return} since $Mac^F$ is a value.

The model for $\mathtt{MacForge}_{A,\widetilde{\Pi}}$, denoted by $\widetilde{\mathit{MacForge}}$, is obtained from the term $\mathit{MacForge}$ from Figure~\ref{fig:macforge-term} instantiating the terms $Mac$, $Gen$ and $Oracle$ according to $\widetilde{\Pi}$, and is represented in Figure~\ref{fig:macforgef-subterms}. The model for $\widetilde{MacForge}$ relies on a truly random function, which is characterized by the property that each distinct input is deterministically associated with a uniformly random output upon its first evaluation, and that this output remains fixed for all subsequent evaluations of the random function on the same input. To model this behavior within $\lamBLL$, we follow the same ledger based approach introduced in~\cite{lago2024-extended}, which is conceptually analogous to the mechanism used to record oracle queries. We introduce an additional ledger that can be manipulated through a collection of function symbols, detailed in Appendix~\ref{app:1}. As shown in Figure~\ref{fig:macforgef-subterms}, this ledger is stored in the reference $rand$, with type $\SS[(q+1)\times(2i+1)]$. The ledger can be viewed as a table with $q+1$ rows, where each row contains an input-output pair of the random function together with a final bit indicating whether the row is active. The size of the ledger reflects the fact that the random function $f$ can be queried at most $q+1$ times: once by the experiment itself and up to $q$ times by the adversary.

\begin{figure}[htpb]
    \begin{align*}
        \begin{aligned}
            &\widetilde{Gen} \triangleq \\
            &\quad \lambda u. \\
            &\quad\letin{q=\mathtt{initQ}}{} \\
            &\quad \mathtt{set}\,Q\,q; \\
            &\quad \letin{tab=\mathtt{initTable}}{} \\
            &\quad \mathtt{set}\,rand\,tab; \\
            &\quad \letin{z=\mathtt{zero}}{} \\
            &\quad \return z
        \end{aligned}
        \qquad\qquad\qquad
        \begin{aligned}
            &\widetilde{\mathit{Mac}} \triangleq \\
            &\quad\lambda e. \\
            &\quad\letin{\lis{k,m} = e}{} \\
            &\quad\letin{tab = \mathtt{get}\,rand}{} \\
            &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
            &\quad\mathtt{if}\, b\, \mathtt{then} \\
            &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
            &\quad\quad \return x \\
            &\quad\mathtt{else} \\
            &\quad\quad \letin{x = \random}{} \\
            &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
            &\quad\quad \mathtt{set}\,rand\,newtab; \\
            &\quad\quad \return x
        \end{aligned}
    \end{align*}
    \begin{align*}
        &\widetilde{Oracle} \triangleq \\
        &\quad \return(\lambda k.\return \bang(\return \lambda m. \\
        &\quad\quad \letin{q=\get Q}{} \\
        &\quad\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad\quad \mathit{\widetilde{Mac}}\lis{k,m}))
    \end{align*}
    \caption{Models for subterms of $\widetilde{MacForge}$.}
    \label{fig:macforgef-subterms}
\end{figure}

The key generation algorithm of $\widetilde{\Pi}$ is modeled by the term $\widetilde{Gen}$. Its role is to initialize the ledger representing the random function by setting the reference $rand$ to the output of the function symbol $\mathtt{initTable}$, which generates a string of length $(q+1)\times(2i+1)$ filled with zeros, corresponding to an empty table. It also initializes the ledger representing the query history by setting the value of the reference $Q$ similarly to $Gen^F$. The term then returns a dummy value generated using the function symbol $\mathtt{zero}$, as the key is not used in the
experiment $\widetilde{MacForge}$.

The term $\widetilde{\mathit{Mac}}$ models the MAC algorithm of $\widetilde{\Pi}$ built from the random function $f$. Given an input message $m$, it samples a fresh random value $r$ and inspects the ledger stored in $rand$ to determine whether $r$ has already been used as an input to the random function. This check is performed using the predicate $\mathtt{isdefined}$ which is reported in Table~\ref{table:function-symbols}. If $r$ is already present in the ledger, the corresponding output value $x$ is retrieved using $\mathtt{getValue}$, and the
resulting tag is computed deterministically. Otherwise, if $r$ is a fresh input, the term samples a new uniformly random value $x$, records the association $(r,x)$ in the ledger using $\mathtt{modify}$, updates the reference $rand$, and then returns the corresponding tag. In this way, the ledger ensures that repeated evaluations of $f$ on the same input yield consistent outputs. 

As in the case of the query ledger, the algorithms of the scheme $\widetilde{\Pi}$ operate directly on the ledger stored in the reference $rand$. Accordingly, the typing of the terms in Figure~\ref{fig:macforgef-subterms} is carried out in a reference context $\Upsilon = \{\, rand : \SS[(q+1)\times(2i+1)], Q: \SS[q\times(i+1)] \,\}$ and the context $\Theta$ appearing in the corresponding typing judgments is instantiated with $\Upsilon$. It is worth observing that the underlying function $f$ can be queried at most $q+1$ times: up to $q$ times by the adversary through its oracle access, and once by the experiment itself during verification via an oracle evaluation on the candidate message generated by the adversary. The size of the ledger stored in $rand$ reflects this bound, as it contains exactly $q+1$ rows, corresponding to the maximum number of distinct messages on which $f$ may be evaluated throughout the execution of the experiment.

In the term $\widetilde{Oracle}$ in Figure~\ref{fig:macforgef-subterms} we use the syntactic sugar of Equation~\ref{eq:sugar-return} since $\widetilde{Mac}$ is a value.

\section{Modeling The Distinguishers}

The distinguisher $D_A$ constructed from any adversary that succeeds in the $\macexp$ experiment, and used in the first step of the proof of Theorem~\ref{theorem:mac-security}, emulates the experiment $\macexp$. The model for the distinguisher is very similar to that of the experiment, however in this case the oracle $o$ is taken as an explicit input. This allows the same term to be used to model both $D_A^{F_k(\cdot)}$ and $D_A^{f(\cdot)}$, namely the distinguisher that receives oracle access to a pseudorandom function $F_k$ and the one that receives oracle access to a truly random function $f$. The $\lamBLL$ term defining the distinguisher is shown in Figure~\ref{fig:d-term}.

\begin{figure}[ht]
    \centering
    \begin{align*}
        &D \triangleq \\
        &\quad \return(\lambda o. \\
        &\quad \letin{\lis{m,tag} = Adv\, o}{} \\
        &\quad \letin{queries = \get Q}{} \\
        &\quad \letin{c=\mathtt{contains}(queries, m)}{} \\
        &\quad \mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\quad \mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag))}
    \end{align*}
    \caption{$\lamBLL$ term for the $\mathtt{D}$ distinguisher.}
    \label{fig:d-term}
\end{figure}

The distinguisher invokes the adversary with oracle access to $o$, obtains a candidate forgery $(m,t)$, and checks its validity using a verification procedure identical to the one used in the $\macexp$ in Figure~\ref{fig:macforge-term}. As in the $\mathtt{MacForge}$ experiment, the reference $Q$ is shared between the oracle and the distinguisher and records all oracle queries performed by the adversary. The distinguisher returns $\true$ if $(m,t)$ is a valid fresh forgery, and
$\false$ otherwise.

We briefly justify that the distinguisher $D$ is well typed in $\lamBLL$. The term $D$ is a value returning a function that takes an oracle $o$ as input which has type $\bang_{q+1}(\SS[i] \multimap \SS[i])$, matching the oracle type expected by the adversary $Adv$. Since $Adv$ is typed as $\bang_{q}(\SS[i] \multimap \SS[i]) \multimap (\SS[i] \otimes \SS[i])$ for a polynomial $q$ the application $Adv\,o$ is well typed and yields a candidate forgery $(m,t)$ with type $(\SS[i] \otimes \SS[i])$. The distinguisher then reads the reference $Q$ using $\get Q$. As in the $\mathtt{MacForge}$ experiment, the oracle $o$ and the distinguisher $D$ operate within the same reference context $\Theta$, which contains the reference $Q : \SS[q\times(i+1)]$. This ensures that the query history recorded by the oracle during the execution of $Adv$ is available to the distinguisher when performing the check for the freshness of the message. It is important to note that modifying the reference state would not correspond to a meaningful strategy for distinguishing a pseudorandom function from a truly random one. The verification step is carried out by checking if the message generated by the adversary belongs to the set of registered queries, if it does not then $D$ evaluates the oracle on the message checking if the tag generated by the adversary and the tag generated by the oracle are equal. The type for the distinguisher follows the type derivation $\vdash_c^\Theta \bang_{q+1}(\SS[i] \multimap \SS[i]) \multimap \BB$

We can now model the distinguishers $D^{F_k(\cdot)}$ and $D^{f(\cdot)}$ as follows:
\begin{align}
D^F &\triangleq \letin{k=Gen^F \star} \letin{o=M^F k} \letin{e=\mathit{MacOracle}\,o}{D\,e} \\  
\widetilde{D} &\triangleq \letin{k=\widetilde{Gen} \star} \letin{o=\widetilde{M} k} \letin{e=\mathit{MacOracle}\,o}{D\,e} \label{eq:d-tilde}
\end{align}

where the subterms $M^F$ and $\widetilde{M}$ defined in Figure~\ref{fig:oracles} model the pseudorandom function $F$ and the random function $f$, respectively.

\begin{figure}[htpb]
\small{
    \centering
    \begin{subfigure}[c]{0.45\textwidth}
        \begin{align*}
        \begin{aligned}
            M^F &\triangleq \\
            &\return(\lambda k.\return(\bang(\return(\lambda m.\\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
            &\qquad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\, \mathtt{skip}; \\
            &\quad \return F\lis{k,m}
        \end{aligned}
        \end{align*}
        \caption{Subterm with pseudorandom function.}
        \label{subfig:oracle-pseudo-random}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.45\textwidth}
        \begin{align*}
        \begin{aligned}
            \widetilde{M} &\triangleq \\
            &\return(\lambda k.\return(\bang(\return(\lambda r. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q, r)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q, r) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \letin{tab=\mathtt{get}\,rand}{} \\
            &\quad \letin{b=\mathtt{isdefined}(tab, r)}{} \\
            &\quad \mathtt{if}\,b\,\mathtt{then} \\
            &\quad\quad \letin{x=\mathtt{getValue}(tab, r)}{} \\
            &\quad\quad \return x \\
            &\quad \mathtt{else} \\
            &\quad\quad \letin{x=\random}{} \\
            &\quad\quad \letin{newtab=\mathtt{modify}(tab, r, x)}{} \\
            &\quad\quad \mathtt{set}\,rand\,newtab; \\
            &\quad\quad \return x))))
        \end{aligned}
        \end{align*}
        \caption{Subterm with truly random function.}
        \label{subfig:oracle-truly-random}
    \end{subfigure}
    \caption{Models for function oracles used by the distinguisher.}
    \label{fig:oracles}
}
\end{figure}

The terms $M^F$ and $\widetilde{M}$ are typed under the following type derivations:
\begin{align*}
&\vdash_c^\Xi M^F: \SS[i] \multimap \bang_{q+1}(\SS[i] \multimap \SS[i]) \\
&\vdash_c^\Upsilon \widetilde{M}: \SS[i] \multimap \bang_{q+1}(\SS[i] \multimap \SS[i])
\end{align*}
These judgments are parameterized by two distinct reference contexts, denoted by $\Xi$ and $\Upsilon$, which correspond to the two different oracles that may be provided to the distinguisher $D$. When the distinguisher is given access to the oracle $M^F$, modeling a pseudorandom function, the reference context $\Theta$ is instantiated as $\Xi = \{\, Q : \SS[q \times (i+1)] \,\}$ where the reference $Q$ is used to record the messages queried to the oracle. In contrast, when the distinguisher is given access to the term $\widetilde{M}$, modeling a truly random function $f$, the reference context $\Theta$ is instantiated as $\Upsilon = \{\, \mathit{rand} : \SS[(q+1) \times (2i+1)],\, Q : \SS[q \times (i+1)] \,\}$. In this case, the additional reference $\mathit{rand}$ is required to store previously generated random outputs which models the truly random function.

The oracle $\mathit{MacOracle}$, defined in Equation~\ref{eq:mac-oracle}, returns a function that is parameterized by an underlying function oracle $o$. In the present setting, $o$ abstracts either a pseudorandom function or a truly random function through terms in Figure~\ref{fig:oracles} and is used to model the MAC computation. Upon receiving a message $m$, the returned function simply invokes the oracle $o$ on $m$ and returns the resulting value, which is interpreted as the corresponding tag.  The term is well typed under the judgment $\vdash_c^\varnothing \mathit{MacOracle}: \bang_{q+1}(\SS[i] \multimap \SS[i]) \multimap \bang_{q+1}(\SS[i] \multimap \SS[i])$ where the graded modality accounts for the single use of the underlying oracle within the returned function.
% TODO: is q+1 good here? or q+2?

\begin{equation}
\begin{aligned}\label{eq:mac-oracle}
    &\mathit{MacOracle} \;\triangleq\;
    \begin{aligned}[t]
        &\return(\lambda o.\return(\bang(\return(\lambda m. \\ 
        &\quad \return(\mathtt{der(o)}\,m) \\
        &))))
    \end{aligned}
\end{aligned}
\end{equation}

% The term $D$ is well typed following the type derivation 
% \begin{align*}
%  \vdash_c^\Theta D:\bang_{q+1}(\SS[i] \multimap \SS[i]) \multimap \BB   
% \end{align*}

\section{Formalization of the Proof}
The security of $\Pi^F$ is established via a proof by reduction (Theorem~\ref{theorem:mac-security}), proceeding contrapositively: from any hypothetical adversary $Adv$ against $\Pi^F$ we obtain a distinguisher $D$ for $F$ which is designed to distinguish the pseudorandom function $F$ from a truly random function $f$. If $D$ is shown to be successful whenever $Adv$ succeeds, we conclude that $\Pi^F$ is secure provided $F$ is a pseudorandom function where security is characterized by the non-existence of efficient adversaries of the appropriate kind.

In the context of ${\lamBLL}$, the distinguisher $D$ is represented as the term defined in Figure~\ref{fig:d-term}, having type $\bang_{q+1}(\mathbb{S}[i] \multimap \mathbb{S}[i]) \multimap \mathbb{B}$. We consider two specific instances of this distinguisher: $D^F$, which interacts with the pseudorandom function $F$, and $\widetilde{D}$, which interacts with a truly random function $f$. Both $F$ and $f$ are modeled using stateful computations. Specifically, a ledger $Q$ records the query history, and in the case of the truly random function, an additional ledger $rand$ is used to provide consistent random outputs. To complete the formalization, we defined an ideal MAC scheme $\widetilde{\Pi}$ that is structurally identical to $\Pi^F$ but uses a truly random function. The components of $\Pi^F$ are specified in Figure~\ref{fig:macforgeF-subterms}, while those of $\widetilde{\Pi}$ are detailed in Figure~\ref{fig:macforgef-subterms}. 

By defining the reference contexts $\Xi=\{Q: \mathbb{S}[q\times(i+1)]\}$ and $\Upsilon = \{rand: \mathbb{S}[(q+1)\times(2i+1)], Q: \mathbb{S}[q\times(i+1)]\}$, the security of $\mathit{MacForge}^F$ is expressed by the equation:
\begin{equation}\label{eq:macforgeF-sim-false}
    \mathit{MacForge}^F \sim_{\Xi\subseteq\Upsilon} \false
\end{equation}
where $\sim_{\Xi\subseteq\Upsilon}$ denotes the $\Theta$-contextual indistinguishability relation as defined in \cite{lago2024computationalindistinguishabilitylogicalrelations}. Note that since $\mathit{MacForge}^F$ is a computation, the term $\false$ appearing on the right-hand side is implicitly lifted to a computation using the syntactic sugar introduced in Equation~\ref{eq:sugar-return}, that is, $\false$ stands for $\return\false$. 

Equation~\ref{eq:macforgeF-sim-false} asserts that the experiment $\mathit{MacForge}^F$ reveals nothing to the adversary, even when accounting for the stateful references stored in the context $\Upsilon$. Although this appears to be a strong requirement, we show that it can be established by structuring the reduction proof through a combination of equational reasoning and reasoning based on Hoare logic with assertion-typed terms. This allows us to derive Equation~\ref{eq:macforgeF-sim-false}, thereby obtaining a formal proof of Theorem~\ref{theorem:mac-security}. The diagram shows that the left component ($\mathit{MacForge}^F$) can be proved insecure only if the lower relation in the diagram is also violated. Since we assume that the pseudorandom function hypothesis holds (see Definition~\ref{def:prf}), it follows that the corresponding MAC is secure.

\begin{figure}[htpb]
	\centering
	\begin{tikzpicture}[xscale=1.6, yscale=1.8]
		\node at (0,0) {$\false$};

		\node[rotate=45]  (A) at (-1,-0.5) {$\not\sim_{\Xi\subseteq\Upsilon}$};
		\node[rotate=-45] at (1,-0.5) {$\sim_{\Xi\subseteq\Upsilon}$};

		\node at (-2,-1) {$\mathit{MacForge}^F$};
		\node at (2,-1) {$\widetilde{\mathit{MacForge}}$};

		\node[rotate=90] at (-2,-1.6) {$\sim_{\Xi\subseteq\Upsilon}$};
		\node[rotate=90] at (2,-1.6) {$\sim_{\Xi\subseteq\Upsilon}$};

		\node at (-2,-2.2) {$D^F$};
		\node at (2,-2.2) {$\widetilde{D}$};

		\node (B) at (0,-2.2) {$\not\sim_{\Xi\subseteq\Upsilon}$};

		\draw[-{Implies},double] (A) to (B);
	\end{tikzpicture}
	\caption{Outline Proof of Security for $\mathtt{MacForge}$.}
	\label{fig:proofscheme}
\end{figure}

Consider the diagram in Figure~\ref{fig:proofscheme}. Our objective is to demonstrate that, on both the left-hand and right-hand sides of the diagram, the relation $\sim_{\Xi \subseteq \Upsilon}$ can be used to relate the corresponding terms. On the left-hand side, we must establish that the contextual equivalence  $D^F \simcontext MacForge^F$ holds. Similarly, on the right-hand side, we must show that $\widetilde{D} \simcontext \widetilde{MacForge}$ holds. These two equivalences correspond respectively to Equation~\ref{eq:macforgeF-to-DF} and Equation~\ref{eq:macforgef-to-Df} in Proof~\ref{proof:handwavy}.

\subsection{From \texorpdfstring{$\widetilde{D}$}{MacOracle} to \texorpdfstring{$\widetilde{MacForge}$}{MacForge}}
We begin by focusing on the bottom-right side of the diagram. The first step is to establish the relation $\letin{o=\widetilde{M}\,k}{MacOracle\,o} \simcontext \widetilde{Oracle}\,k$, which is stated as Lemma~\ref{lemma:MacOracle-to-tilde-oracle} and proved immediately after. Once this equivalence is established, the relation $\widetilde{D} \simcontext \widetilde{MacForge}$ follows by an application of Equation~\ref{letAss} and Lemma~\ref{lemma:MacOracle-to-tilde-oracle}, and by concluding with a reduction using Equation~\ref{reduc2}.

\begin{lemma}\label{lemma:MacOracle-to-tilde-oracle}
\[ \letin{o=\widetilde{M}\,k}{MacOracle\,o} \simcontext \widetilde{Oracle}\,k \]
\end{lemma}
\begin{proof}
We begin by unfolding the definition of $\widetilde{Oracle}$ and applying Equation~\ref{reduc2}. We then unfold the definition of $\widetilde{Mac}$, after which the proof is completed by straightforward applications of Equations~\ref{reduc4}, \ref{reduc3}, and finally \ref{reduc2}. Observe that the equations used, which are reported in Figure~\ref{fig:equations}, are parameterized by two contexts, $\Theta$ and $\Xi$. In this case, $\Theta = \Xi = \{\, Q : \mathbb{S}[q \times (i+1)] \,\}$ and $\Xi = \Upsilon = \{\, rand : \mathbb{S}[(q+1) \times (2i+1)], Q : \mathbb{S}[q \times (i+1)]\,\}$, since both the references $Q$ and $rand$ are accessed during the reduction.

\small{
\begin{align}
    &\widetilde{Oracle}\,k \quad\triangleq\quad
    \begin{aligned}
        &\quad \return(\lambda k.\return \bang(\return \lambda m. \\
        &\qquad \letin{q=\get Q}{} \\
        &\qquad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\qquad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\qquad \mathtt{else}\ \mathtt{skip}; \\
        &\qquad \mathit{\widetilde{Mac}}\lis{k,m} \\
        &\quad))\,k
    \end{aligned} \nonumber
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
        &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m)\\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad \mathit{\widetilde{Mac}}\lis{k,m} \\
        &)))
    \end{aligned} \nonumber
    \quad\triangleq\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
        &\qquad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m)\\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad (\lambda e. \\
        &\quad\quad\letin{\lis{k,m} = e}{} \\
        &\quad\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad\quad \return x \\
        &\quad\quad\mathtt{else} \\
        &\quad\quad\quad \letin{x = \random}{} \\
        &\quad\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad\quad \return x \\
        &\quad)\lis{k,m})))
    \end{aligned} \nonumber
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then}\, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m)\\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad\letin{\lis{k,m} = \lis{k,m}}{} \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad \return x \\
        &\quad\mathtt{else} \\
        &\quad\quad \letin{x = \random}{} \\
        &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad \return x \\
        &)))
    \end{aligned} \nonumber
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc5}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then}\, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m)\\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad \return x \\
        &\quad\mathtt{else} \\
        &\quad\quad \letin{x = \random}{} \\
        &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad \return x \\
        &)))
    \end{aligned} \label{eq:reduced-oracle}
\end{align}
}

The remaining goal is to show that the term $\letin{o=\widetilde{M}\,k}{MacOracle\,o}$ is $\Theta$-contextually indistinguishable from the term shown in Equation~\ref{eq:reduced-oracle}. We begin by unfolding the definition of $\mathit{MacOracle}$, as given in Equation~\ref{eq:mac-oracle}. This term is a curried function that, when applied to an oracle $o$, returns the corresponding MAC computation. We then apply Equation~\ref{reduc2} to eliminate the lambda abstraction and simplify the application of $\mathit{MacOracle}$ to $o$. Next, we unfold the definition of $\widetilde{M}$ from Figure~\ref{subfig:oracle-truly-random}. Applying $\widetilde{M}$ to a key $k$ yields a function that, on input $m$, inspects the ledger to determine whether a value has already been associated with $m$. If so, the stored value is returned; otherwise, fresh randomness is sampled, the reference storing the ledger is updated, and the newly sampled value is returned. We then apply Equation~\ref{reduc2} to evaluate this application. Finally, the term is simplified by successive applications of Equations~\ref{reduc4}, \ref{reduc3}, and \ref{reduc2}.

\small{
    \begin{align*}
        &\letin{o=\widetilde{M}\,k}{MacOracle\,o} \quad\triangleq\quad
        \begin{aligned}
            &\letin{o=\widetilde{M}\,k}{} \\
            &\return(\lambda o.\return(\bang(\return(\lambda m. \\ 
            &\quad \return(\mathtt{der(o)}\,m) \\
            &))))\,o
        \end{aligned}
        \\[2ex]
        &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
        \begin{aligned}
            &\letin{o=\widetilde{M}\,k}{} \\
            &\return(\bang(\return(\lambda m. \\ 
            &\quad\return(\mathtt{der(o)}\,m) \\
            &)))
        \end{aligned}
        \quad\triangleq\quad
        \begin{aligned}
            &\mathtt{let}\,o=\left(
                \begin{aligned}
                    &\return(\lambda k.\return(\bang(\return(\lambda r. \\
                    &\quad \letin{q=\get Q}{} \\
                    &\quad \letin{t=\mathtt{contains}(q, r)}{} \\
                    &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
                    &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, r) \\
                    &\quad \mathtt{else}\ \mathtt{skip}; \\
                    &\quad \letin{tab=\mathtt{get}\,rand}{} \\
                    &\quad \letin{b=\mathtt{isdefined}(tab, r)}{} \\
                    &\quad \mathtt{if}\,b\,\mathtt{then} \\
                    &\quad\quad \letin{x=\mathtt{getValue}(tab, r)}{} \\
                    &\quad\quad \return x \\
                    &\quad \mathtt{else} \\
                    &\quad\quad \letin{x=\random}{} \\
                    &\quad\quad \letin{newtab=\mathtt{modify}(tab, r, x)}{} \\
                    &\quad\quad \mathtt{set}\,rand\,newtab; \\
                    &\quad\quad \return x \\
                    &))))
                \end{aligned}
            \right) k\,\mathtt{in} \\
            &\return(\bang(\return(\lambda m.\return(\mathtt{der(o)}\,m)))
        \end{aligned}
        \\[2ex]
        &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
        \begin{aligned}
            &\mathtt{let}\,o=\left(
            \begin{aligned}
                &\return(\bang(\return(\lambda r. \\
                &\quad \letin{q=\get Q}{} \\
                &\quad \letin{t=\mathtt{contains}(q, r)}{} \\
                &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q, r) \\
                &\quad \mathtt{else}\ \mathtt{skip}; \\
                &\quad \letin{tab=\mathtt{get}\,rand}{} \\
                &\quad \letin{b=\mathtt{isdefined}(tab, r)}{} \\
                &\quad \mathtt{if}\,b\,\mathtt{then} \\
                &\quad\quad \letin{x=\mathtt{getValue}(tab, r)}{} \\
                &\quad\quad \return x \\
                &\quad \mathtt{else} \\
                &\quad\quad \letin{x=\random}{} \\
                &\quad\quad \letin{newtab=\mathtt{modify}(tab, r, x)}{} \\
                &\quad\quad \mathtt{set}\,rand\,newtab; \\
                &\quad\quad \return x \\
                &)))
            \end{aligned}
            \right) \mathtt{in} \\
            &\return(\bang(\return(\lambda m.\return(\mathtt{der(o)}\,m)))
        \end{aligned}
        \\[2ex]    
        &\quad\overset{\text{~\ref{reduc4}}}{\simcontext}\quad
        \begin{aligned}
            &\return(\bang(\return(\lambda m. \\ 
            &\quad\return\left(\mathtt{der}\left(\bang\left(
            \begin{aligned}
                &\return(\lambda r. \\
                &\quad \letin{q=\get Q}{} \\
                &\quad \letin{t=\mathtt{contains}(q, r)}{} \\
                &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q, r) \\
                &\quad \mathtt{else}\, \mathtt{skip}; \\
                &\quad \letin{tab=\mathtt{get}\,rand}{} \\
                &\quad \letin{b=\mathtt{isdefined}(tab, r)}{} \\
                &\quad \mathtt{if}\,b\,\mathtt{then} \\
                &\quad\quad \letin{x=\mathtt{getValue}(tab, r)}{} \\
                &\quad\quad \return x \\
                &\quad \mathtt{else} \\
                &\quad\quad \letin{x=\random}{} \\
                &\quad\quad \letin{newtab=\mathtt{modify}(tab, r, x)}{} \\
                &\quad\quad \mathtt{set}\,rand\,newtab; \\
                &\quad\quad \return x)
            \end{aligned}
            \right)\right)\,m\,\right) \\
            &)))
        \end{aligned}
        \\[2ex]
        &\quad\overset{\text{~\ref{reduc3}}}{\simcontext}\quad
        \begin{aligned}
            &\return(\bang(\return(\lambda m. \\ 
            &\quad\return \left(
            \begin{aligned}
                &\return(\lambda r. \\
                &\quad \letin{q=\get Q}{} \\
                &\quad \letin{t=\mathtt{contains}(q, r)}{} \\
                &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
                &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, r) \\
                &\quad \mathtt{else}\ \mathtt{skip}; \\
                &\quad \letin{tab=\mathtt{get}\,rand}{} \\
                &\quad \letin{b=\mathtt{isdefined}(tab, r)}{} \\
                &\quad \mathtt{if}\,b\,\mathtt{then} \\
                &\quad\quad \letin{x=\mathtt{getValue}(tab, r)}{} \\
                &\quad\quad \return x \\
                &\quad \mathtt{else} \\
                &\quad\quad \letin{x=\random}{} \\
                &\quad\quad \letin{newtab=\mathtt{modify}(tab, r, x)}{} \\
                &\quad\quad \mathtt{set}\,rand\,newtab; \\
                &\quad\quad \return x)
            \end{aligned}
            \right)\, m \\
            &)))
        \end{aligned}
        \\[2ex]
        &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
        \begin{aligned}
            &\return(\bang(\return(\lambda m. \\ 
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q, m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \letin{tab=\mathtt{get}\,rand}{} \\
            &\quad \letin{b=\mathtt{isdefined}(tab, m)}{} \\
            &\quad \mathtt{if}\,b\,\mathtt{then} \\
            &\quad\quad \letin{x=\mathtt{getValue}(tab, m)}{} \\
            &\quad\quad \return x \\
            &\quad \mathtt{else} \\
            &\quad\quad \letin{x=\random}{} \\
            &\quad\quad \letin{newtab=\mathtt{modify}(tab, m, x)}{} \\
            &\quad\quad \mathtt{set}\,rand\,newtab; \\
            &\quad\quad \return x \\
            &)))
        \end{aligned}
        \quad=\quad
        {\text{Term in Equation~\ref{eq:reduced-oracle}}}
    \end{align*}
}
\end{proof}

\begin{theorem}\label{theorem:tildeD-sim-tildeMACForge}
\[ \widetilde{D} \simcontext \widetilde{MacForge} \]
\end{theorem}

\begin{proof}
We begin with the definition of $\widetilde{D}$ given in Equation~\ref{eq:d-tilde}. It describes a computation in which a key $k$ is generated by $\widetilde{Gen}$, after which the function oracle $\widetilde{M}$ is applied to $k$ to produce an oracle $o$. This oracle is then passed to the term $\mathit{MacOracle}$ to produce a MAC oracle $e$, which is finally given as input to the distinguisher $D$ to obtain a boolean outcome. We first apply Equation~\ref{letAss} to restructure the expression which allows us to isolate the application of $\mathit{MacOracle}$. Next, we replace the term $\widetilde{M}$ with $\widetilde{Oracle}$, since these two terms are $\Theta$-contextually indistinguishable by Lemma~\ref{lemma:MacOracle-to-tilde-oracle}. After this, we apply $\alpha$-equivalence, as recalled in Chapter~\ref{chapter:lambda-calculus}, to rename bound variables and align the resulting term with the expected structure of the distinguisher. We then unfold the definition of $D$ from Figure~\ref{fig:d-term} and finally, we apply Equation~\ref{reduc2} to eliminate redundant bindings and Equation~\ref{letCom} to reorder computations, yielding a term that is syntactically equal to $\widetilde{MacForge}$, which concludes the proof.

\small{
\begin{align*}
    \widetilde{D}\triangleq\,& \letin{k=\widetilde{Gen} \star} \letin{o=\widetilde{M} k} \letin{e=\mathit{MacOracle}\,o}{D\,e}      
    \\[2ex]
    \quad \overset{\text{~\ref{letAss}}}{\simcontext}\quad&
    \begin{aligned}
        &\letin{k=\widetilde{Gen} \star}\letin{e=(\letin{o=\widetilde{M}\,k}{MacOracle}\,o)}{D\,e}
    \end{aligned}
    \\[2ex]
    \quad \overset{\text{Lemma ~\ref{lemma:MacOracle-to-tilde-oracle}}} {\simcontext} \quad&
    \begin{aligned}
        &\letin{k=\widetilde{Gen}\star} \\&\letin{e=\widetilde{Oracle}\,k}{} 
        \\&{D\,e}
    \end{aligned}
    \\[2ex]
    \quad=_{\alpha}\quad&
    \begin{aligned}
        &\letin{k=\widetilde{Gen}\star} \\
        &\letin{o=\widetilde{Oracle}\,k}{} \\
        &{D\,o}
    \end{aligned}
    \\[2ex]
    \quad\triangleq\quad&
    \begin{aligned}
        &\letin{k=\widetilde{Gen}\star}\letin{o=\widetilde{Oracle}\,k}{} \\
        &\left(
        \begin{aligned}
            &\return(\lambda o. \\
            &\quad \letin{\lis{m,tag} = Adv\, o}{} \\
            &\quad \letin{queries = \get Q}{} \\
            &\quad \letin{c=\mathtt{contains}(queries, m)}{} \\
            &\quad \mathtt{if}\, c\, \mathtt{then}\, \return \false \\
            &\quad \mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag))}
        \end{aligned}
        \right) o
    \end{aligned}
    \\[2ex]
    \quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad&
    \begin{aligned}
        &\letin{k=\widetilde{Gen}\star}{} \\
        &\letin{o=\widetilde{Oracle}\,k} {} \\
        &\letin{\lis{m,tag} = Adv\, o}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c \,\mathtt{then}\, \return \false \\
        &\mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag))}
    \end{aligned}
    \quad=\quad
    \mathit{\widetilde{MacForge}}
\end{align*}
}
\end{proof}

\subsection{From \texorpdfstring{$D^F$}{D} to \texorpdfstring{$MacForge^F$}{MacForge}}

For the left side of the diagram, we first establish the relation $M^F \simcontext Oracle^F$ in Lemma~\ref{lemma:MacOracle-to-oracleF}. Then, we prove the desired relation $D^F \simcontext MacForge^F$ in Theorem~\ref{theorem:DF-sim-MacForgeF}. The proofs of these statements follow the same structure and reasoning as those presented for the right side of the diagram.

\begin{lemma}\label{lemma:MacOracle-to-oracleF}
\[ \letin{o=M^F}{MacOracle\,o}\simcontext Oracle^F\,k \]
\end{lemma}
\begin{proof}
The proof follows a similar approach to the proof of Lemma~\ref{lemma:MacOracle-to-tilde-oracle}.

\small{
\begin{align}
    &Oracle^F\,k \nonumber \quad\triangleq\quad 
    \begin{aligned}
        &\return(\lambda k.\return \bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad\mathtt{else}\ \mathtt{skip}; \\
        &\quad \mathit{Mac}^F\lis{k,m} \\
        &)))\,k
    \end{aligned} \nonumber
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad\mathtt{else}\ \mathtt{skip}; \\
        &\quad \mathit{Mac}^F\lis{k,m} \\
        &)))
    \end{aligned} \nonumber
    \quad\triangleq\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad\mathtt{else}\ \mathtt{skip}; \\
        &\quad(\lambda x. \\
        &\quad\quad \letin{\lis{k,m}=x}{} \\
        &\quad\quad \return F\lis{k,m} \\
        &\quad)\lis{k,m} \\
        &)))
    \end{aligned} \nonumber
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
        &\qquad\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad\mathtt{else}\ \mathtt{skip}; \\
        &\quad \letin{\lis{k,m}=\lis{k,m}}{} \\
        &\quad \return F\lis{k,m} \\
        &)))
    \end{aligned} \nonumber
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc5}}}{\simcontext}\quad
    \begin{aligned} 
        &\return(\bang(\return(\lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
        &\qquad\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad \return F\lis{k,m} \\
        &)))
    \end{aligned} \label{eq:oracleF-expansion} 
\end{align}
}
Now, the goal is to show that the term $\letin{o=M^F\,k}{MacOracle\,o}$ is $\Theta$-contextual indistinguishable from the term above in Equation~\ref{eq:oracleF-expansion}.

\small{
\begin{align*}
    &\begin{aligned}
    \letin{o=M^F\,k}{MacOracle\,o}
    \end{aligned}
    \quad\triangleq\quad
    \begin{aligned}
        & \letin{o=M^F\,k} \\
        &\return(\lambda o.\return(\bang(\return(\lambda m. \\ 
        &\quad \return(\mathtt{der(o)}\,m) \\
        & ))))\,o
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        & \letin{o=M^F\,k} \\
        &\return(\bang(\return(\lambda m. \\ 
        &\quad \return(\mathtt{der(o)}\,m) \\
        & ))) 
    \end{aligned}
    \quad\triangleq\quad
    \begin{aligned}
        &\mathtt{let}\,o=\left(
        \begin{aligned}
            &\return(\lambda k.\return(\bang(\return(\lambda m.\\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
            &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \return F\lis{k,m} \\
            &))))
        \end{aligned}
        \right)\,k\,\mathtt{in} \\
        &\return(\bang(\return(\lambda m. \\ 
        &\quad \return(\mathtt{der(o)}\,m) \\
        & )))
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\mathtt{let}\,o=\left(
        \begin{aligned}
            &\return(\bang(\return(\lambda m.\\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
            &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \return F\lis{k,m} \\
            &)))
        \end{aligned}
        \right) \mathtt{in} \\
        &\return(\bang(\return(\lambda m. \\ 
        &\quad \return(\mathtt{der(o)}\,m) \\
        & )))
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc4}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\ 
        &\quad \return\left(\mathtt{der}\left(\bang\left(
        \begin{aligned}
            &\return(\lambda m.\\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
            &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \return F\lis{k,m})
        \end{aligned}
        \right)\right)m\right) \\
        & )))
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc3}}}{\simcontext}\quad
    \begin{aligned}
        &\quad \return(\bang(\return(\lambda m. \\ 
        &\quad \left(
        \begin{aligned}
            &\return(\lambda m.\\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
            &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \return F\lis{k,m}
        \end{aligned}
        \right) m \\
        & )))
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\return(\bang(\return(\lambda m. \\ 
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\ 
        &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad \return F\lis{k,m} \\
        &)))
    \end{aligned}
    \quad=\quad
    \text{Term in Equation~\ref{eq:oracleF-expansion}}
\end{align*}
}
\end{proof}

\begin{theorem}\label{theorem:DF-sim-MacForgeF}
\[
D^F \simcontext \mathit{MacForge}^F
\]
\end{theorem}
\begin{proof}
The proof follows a similar approach to the proof of Theorem~\ref{theorem:tildeD-sim-tildeMACForge}.
\small{
\begin{align*}
    D^F \quad\triangleq\quad& \letin{k=Gen^F \star} \letin{o=M^F k}\letin{e=\mathit{MacOracle}\,o}{D\,e}
    \\[2ex]
    \quad\overset{\text{~\ref{letAss}}}{\simcontext}\quad&
    \begin{aligned}
        &\letin{k=Gen^F \star}{\letin{e=(\letin{o=M^F\,k}{MacOracle\,o)}}{D\,e}}
    \end{aligned}
    \\[2ex]
     \quad\overset{\text{Lemma ~\ref{lemma:MacOracle-to-oracleF}}}{\simcontext}\quad&
    \begin{aligned} 
        &\letin{k=Gen^F \star}{} \\
        &\letin{e=Oracle^F\,k}{} \\
        &{D\,e}
    \end{aligned}
    \\[2ex]
    \quad=_\alpha\quad&
    \begin{aligned}
        &\letin{k=Gen^F \star}{} \\
        &\letin{o=Oracle^F\,k}{} \\
        &{D\,o}
    \end{aligned}
    \\[2ex]
    \quad\triangleq\quad&
    \begin{aligned}
        &\letin{k=Gen^F \star}{} \\
        &\letin{o=Oracle^F\,k}{} \\
        &\left(
        \begin{aligned}
            &\return(\lambda o. \\
            &\quad\letin{\lis{m,tag} = Adv\, o}{} \\
            &\quad\letin{queries = \get Q}{} \\
            &\quad\letin{c=\mathtt{contains}(queries, m)}{} \\
            &\quad\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
            &\quad\mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag)})
        \end{aligned}
        \right) o
    \end{aligned}
    \\[2ex]
    \quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad&
    \begin{aligned}
        &\letin{k=Gen^F \star}{} \\
        &\letin{o=Oracle^F\,k}{} \\
        &\letin{\lis{m,tag} = Adv\, o}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag)}
    \end{aligned}
    \quad=\quad\mathit{MacForge^F}
\end{align*}
}
\end{proof}

\subsection{From \texorpdfstring{$\widetilde{MacForge}$}{MacForge} to \texorpdfstring{$\false$}{false}}

To complete the overall security proof, it remains to establish the relation $\widetilde{MacForge}\simcontext \false$, which corresponds to the right side of the diagram in Figure~\ref{fig:proofscheme}.

We introduce assertions, in the style of standard Hoare proofs, which are indicated in blue. The rules are composed using Rule~\ref{rule:h-let} and correspond to derivation trees that establish the propositions annotated after each line. We state two lemmas, namely, Lemma~\ref{lemma:gen-coherence} and Lemma~\ref{lemma:oracle-coherence}, that are used to decorate the term $\widetilde{MacForge}$ and to show that a given assertion holds at a specific point in the term. Once the assertion is established, we use it to simplify part of the term using equations introduced in Section~\ref{sec:extensions-lambda-bll}. Finally, we conclude by standard equational reasoning of $\lamBLL$, namely $\Theta$-contextual indistinguishability, which is applied to the term obtained after the rewriting through Hoare logic.

% \begin{lemma}\label{lemma:coherent-empty-q}
% For any ledger $rand$ and for any ledger $Q$ the predicates $\mathtt{coherent}(\mathtt{initQ}, rand)$ and $\mathtt{coherent}(Q, \mathtt{initTable})$ are true.
% \end{lemma}
% \begin{proof}
% Let $s=\mathtt{initQ}$. From Table~\ref{table:new-function-symbols}, $\mathtt{initQ}$ returns a string of length $q\times(i+1)$ for a polynomial $q$, filled with zeros. In the query ledger which is modelled by $s$ a row is active if and only if the $(i+1)$-th bit is set to 1. Since every bit in $s$ is 0, the activation bit for every row $k\in\{1,\dots,q\}$ is 0. Therefore, there are no active rows in $s$. The predicate $\mathtt{coherent}$ requires that for every active string $s$ in the query ledger, there exists a corresponding entry in the random ledger ($rand$). Since the set of active strings in $s$ is empty ($\emptyset$), the universal quantification is true. The same reasoning can be applied to predicate $\mathtt{coherent}(Q, \mathtt{initTable})$.
% \end{proof}

\begin{lemma}\label{lemma:gen-coherence}
The following Hoare triple holds:
\begin{align*}
\{\top\}\,\widetilde{Gen}\, \star\,\{\lambda \_.\mathtt{coherent}(Q, rand)\}
\end{align*}
\end{lemma}

\begin{proof}
Since we evaluate $\widetilde{Gen}$ at $\star$, we first eliminate the abstraction using \ref{rule:h-lam} and \ref{rule:h-app}. Then we reason about the body of $\widetilde{Gen}$ and we need to prove the triple for the body. Sequential composition is justified by repeated applications of \ref{rule:h-let}.
\small{
\begin{align*}
\begin{aligned}
    &\color{blue} \{\top\} \\
    &\letin{q = \mathtt{initQ}}{} \\
    &\color{blue} \{ \mathtt{coherent}(q, rand) \} \quad \text{\ref{rule:h-fun}, \ref{rule:h-cons}} \\
    &\mathtt{set}\, Q\, q; \\
    &\color{blue} \{ \mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-set}, \ref{rule:h-cons}} \\
    &\letin{tab = \mathtt{initTable}}{} \\
    &\color{blue} \{ \mathtt{coherent}(Q, tab)\} \quad \text{\ref{rule:h-fun}, \ref{rule:h-cons}} \\
    &\mathtt{set}\, rand\, tab; \\
    &\color{blue} \{ \mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-set}} \\
    &\letin{z = \mathtt{zero}}{} \\
    &\color{blue} \{ \mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-fun}} \\
    &\return z \\
    &\color{blue} \{ \lambda z. \mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-eta}}
\end{aligned}
\end{align*}
}
\end{proof}

\begin{lemma}\label{lemma:oracle-coherence}
Let $M$ be the term 
\begin{align*}
M \quad\triangleq\quad
\begin{aligned}
    &\bang (\return \lambda m. \\
    &\quad \letin{q=\get Q}{} \\
    &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
    &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
    &\quad\quad \letin{q'=\mathtt{modifyQ}(q,m)}{} \\
    &\quad\quad \mathtt{set}\,Q\, q' \\
    &\quad\mathtt{else}\ \mathtt{skip}{}; \\
    &\quad\letin{tab = \mathtt{get}\,rand}{} \\
    &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    &\quad\mathtt{if}\, b\, \mathtt{then} \\
    &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
    &\quad\quad \return x \\
    &\quad\mathtt{else} \\
    &\quad\quad \letin{x = \random}{} \\
    &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
    &\quad\quad \mathtt{set}\,rand\,newtab; \\
    &\quad\quad \return x)
\end{aligned}
\end{align*}

The following Hoare triple holds:
\begin{align*}
\{\mathtt{coherent}(Q, rand)\}\,M\,\{\lambda v.\mathtt{coherent}(Q, rand)\}
\end{align*}
\end{lemma}

\begin{proof}
The proof follows the same structure as that of Lemma~\ref{lemma:gen-coherence}. We extend the derivation with assertions justified by the rules introduced in Figure~\ref{fig:hl-typing-rules}. Sequential composition is handled by repeated applications of Rule~\ref{rule:h-let}, while Rule~\ref{rule:h-fun} is applied to function symbols. We reason on the body of the term, implicitly applying Rule~\ref{rule:h-bang}, followed by Rule~\ref{rule:h-eta}, and finally Rule~\ref{rule:h-lam} to eliminate the abstraction. The states of the ledgers $Q$ and $rand$ are described using standard set notation. For instance, $Q \setminus m$ denotes the ledger obtained from $Q$ by removing the element $m$, more formally $Q\setminus\{m\} := \{\, x\in Q \mid x\not= m \,\}$. The ledger $rand$ is specified in assertions as a set of pairs. The notation $rand \cup \{(m,x)\}$ denotes the ledger obtained by adding a new pair $(m,x)$, where $m$ and $x$ are bitstrings of length $\SS[i]$. In the concrete representation, such a pair is modeled as a single entry of type $\SS[2i+1]$, where the additional bit indicates whether the entry is active. The status bit is handled by the underlying logic of the function symbol $\mathtt{modify}$. We use syntactic sugar of Equation~\ref{eq:sugar-letvoid} and ~\ref{eq:sugar-skip} to avoid unnecessary binding and to simplify computations that do not add effects.

\scriptsize{
\begin{align*}
\begin{aligned}
    & \color{blue} \{ \mathtt{coherent}(Q, rand) \} \\
    & \letin{q = \get Q}{} \\
    & \color{blue} \{ \mathtt{coherent}(q,rand) \land (q = Q) \} \quad \text{\ref{rule:h-get}} \\
    & \letin{t = \mathtt{contains}(q, m)}{} \\
    & \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land t=(m \in Q) \} \quad \text{\ref{rule:h-fun}, \ref{rule:h-cons}} \\
    & \mathtt{if}\ \mathtt{not}(t)\ \mathtt{then} \\
    & \quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (q = Q) \land t=(m \in Q) \land (t = \false) \} \quad \text{\ref{rule:h-case}} \\
    & \qquad \implies \color{blue} \{ \mathtt{coherent}(q, rand) \land (q = Q) \land (m \notin Q) \} \quad \text{\ref{rule:h-cons}} \\
    & \quad \letin{q' = \mathtt{modifyQ}(q, m)}{} \\
    & \quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \notin Q) \land q' = \mathtt{modifyQ}(q, m) \} \quad \text{\ref{rule:h-fun}} \\
    &\qquad \implies \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \notin Q) \land q' = (q \cup m) \} \quad \text{\ref{rule:h-cons}} \\
    &\quad\qquad \implies \color{blue} \{ \mathtt{coherent}(q' \setminus \{m\}, rand) \land (q=Q) \land (m \notin Q) \land q' = (q \cup m) \} \quad \text{\ref{rule:h-cons}} \\
    & \quad \mathtt{set}\ Q\ q' \\
    & \quad \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, rand) \land (m \in Q) \} \quad \text{\ref{rule:h-set}} \\
    & \mathtt{else} \\
    & \quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \in Q) \} \quad \text{\ref{rule:h-case}, \ref{rule:h-cons}} \\
    & \quad \mathtt{skip}; \\
    & \quad \color{blue} \{\mathtt{coherent}(q, rand) \land (q=Q) \land (m \in Q) \} \quad \text{\ref{rule:h-eta}} \\
    & \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \in Q) \} \implies \\
    & \color{blue} \quad \{ \mathtt{coherent}(Q \setminus \{m\}, rand) \land (m \in Q) \} \quad \text{\ref{rule:h-cons}} \\
    &\letin{tab = \mathtt{get}\,rand}{} \\
    & \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab = rand) \} \quad \text{\ref{rule:h-get}} \\
    &\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    & \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab = rand) \land b = (m \in tab) \} \quad \text{\ref{rule:h-fun}} \\
    &\mathtt{if}\, b\, \mathtt{then} \\
    & \quad \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab=rand) \land b=(m \in tab) \land (b=\true) \} \quad \text{\ref{rule:h-case}} \\
    & \qquad \implies \color{blue} \{ \mathtt{coherent}(Q, tab) \land (tab=rand) \} \quad \text{\ref{rule:h-cons}} \\
    & \quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
    &\quad \return x \\
    & \color{blue} \quad \{ \lambda x.\mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-cons}, \ref{rule:h-fun}, \ref{rule:h-eta}} \\
    &\mathtt{else} \\
    & \quad \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab=rand) \land b=(m \in tab) \land (b=\false) \} \quad \text{\ref{rule:h-case}} \\
    & \qquad \implies \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab=rand) \land (m \notin tab) \} \quad \text{\ref{rule:h-cons}} \\
    &\quad \letin{x = \random}{} \\
    &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
    & \quad \color{blue}\{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab=rand) \land (m \notin tab) \land newtab = \mathtt{modify}(tab, m, x) \} \quad \text{\ref{rule:h-fun}} \\
    & \qquad \implies \color{blue}\{ \mathtt{coherent}(Q \setminus \{m\}, tab) \land (m \in Q) \land (tab=rand) \land (m \notin tab) \land newtab = (tab \cup \{(m, x)\}) \} \quad \text{\ref{rule:h-cons}} \\
    & \quad\qquad \implies \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, tab \cup \{ (m, x) \}) \land (m \in Q) \land (tab=rand) \land (m \notin tab) \land newtab = (tab \cup \{ (m, x) \} \} \quad \text{\ref{rule:h-cons}} \\
    &\quad \mathtt{set}\,rand\,newtab; \\
    &\quad \color{blue} \{ \mathtt{coherent}(Q \setminus \{m\}, rand) \land (m \in Q) \land (m \in rand) \} \implies \color{blue} \{ \mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-set}, \ref{rule:h-cons}} \\
    &\quad \return x \\
    &\quad \color{blue} \{ \lambda x.\mathtt{coherent}(Q, rand) \} \quad \text{\ref{rule:h-eta}}
\end{aligned}
\end{align*}
}
\end{proof}

\normalsize

% \begin{mathpar} \mprset{sep=1em, vskip =-0.1ex} \scriptsize 
% \inferrule* [left=\scriptsize $\mathcal{D}_{initQ}$, right=\scriptsize \text{h-cons}]
% {
%     \inferrule* [right=\scriptsize \text{h-fun}]
%     {
%         \mathsf{typeof}(\mathtt{initQ}) = \mathbb{S}[p] \\
%         p \in \mathbb{N}_{\geq 1}[i]
%     }
%     { \vdashtheta{c} \{ \mathtt{coherent}(\mathtt{initQ}, rand) \} \, \mathtt{initQ} \, \{ \lambda q. \mathtt{coherent}(q, rand) \} : \SS }
%     \and 
%     \top \vdash \mathtt{coherent}(\mathtt{initQ}, rand)
% }
% { \vdashtheta{c} \{ \top \} \, \mathtt{initQ} \, \{ \lambda q. \mathtt{coherent}(q, rand) \} }
% \and
% \inferrule* [left=\scriptsize $\mathcal{D}_{setQ}$, right=\scriptsize \text{h-set}]
% {
%     q : \mathbb{S}[p] \vdashtheta{v} q : \mathbb{S}[p] \\
%     Q : \mathbb{S}[p] \in \Theta
% }
% { q : \mathbb{S}[p] \vdashtheta{c} \{ \mathtt{coherent}(q, rand) \} \, \setr{Q}{q} \, \{ \lambda \_. \mathtt{coherent}(Q, rand) \} }
% \and
% \inferrule* [right=\scriptsize \text{h-let}]
% {
%     \mathcal{D}_{initQ}
%     \and 
%     \mathcal{D}_{setQ}
% }
% { \vdashtheta{c} \{ \top \} \, \letin{q = \mathtt{initQ}}{\setr{Q}{q}} \, \{ \lambda \_. \mathtt{coherent}(Q, rand) \} : \mathbb{U} }
% \and
% \inferrule* [left=\scriptsize $\mathcal{D}_{initRand}$, right=\scriptsize \text{h-cons}]
% {
%     \inferrule* [right=\scriptsize \text{h-fun}]
%     {
%         \mathsf{typeof}(\mathtt{initTable}) = \mathbb{S}[p] \\
%         p \in \mathbb{N}_{\geq 1}[i]
%     }
%     { \vdashtheta{c} \{ \mathtt{coherent}(\mathtt{initTable}, rand) \} \, \mathtt{initTable} \, \{ \lambda q. \mathtt{coherent}(q, rand) \} : \SS }
%     \and 
%     \mathtt{coherent}(Q, \mathtt{initTable}) \vdash \top
% }
% { \vdashtheta{c} \{ \mathtt{coherent}(Q, rand) \} \, \mathtt{initTable} \, \{ \lambda t. \mathtt{coherent}(Q, t) \} }
% \and
% \inferrule* [left=\scriptsize $\mathcal{D}_{setRand}$, right=\scriptsize \text{h-set}]
% {
%     q : \mathbb{S}[p] \vdashtheta{v} q : \mathbb{S}[p] \\
%     Q : \mathbb{S}[p] \in \Theta
% }
% { tab : \mathbb{S}[p] \vdashtheta{c} \{ \mathtt{coherent}(q, rand) \} \, \setr{Q}{q} \, \{ \lambda \_. \mathtt{coherent}(Q, rand) \} }
% \and
% \inferrule* [right=\scriptsize \text{h-let}]
% {
%     \mathcal{D}_{initRand}
%     \and 
%     \mathcal{D}_{setRand}
% }
% { \vdashtheta{c} \{ \mathtt{coherent}(Q, rand) \} \, \letin{tab = \mathtt{initTable}}{\setr{rand}{tab}} \, \{ \lambda \_. \mathtt{coherent}(Q, rand) \} : \mathbb{U} }
% \end{mathpar}

% \begin{center} \scriptsize
% \begin{mathpar}\mprset{sep=1em,  vskip =-0.1ex}
% \inferrule* [right=\footnotesize h-let]
% {
%     \inferrule*[right=\scriptsize h-cons]
%     {
%         \inferrule*[right=\scriptsize h-eta]
%         { \Gamma \vdashtheta{v} \mathtt{initQ} : \SS }
%         { \Gamma \vdashtheta{c} \{ \top \} \, \return \mathtt{initQ} \, \{ \lambda v. v = \mathtt{initQ} \} : \SS }
%         \\
%         v = \mathtt{initQ} \vdash v = \emptyset
%     }
%     { \Gamma \vdashtheta{c} \{ \top \}\, \mathtt{initQ} \,\{\lambda v. v = \emptyset\}: \SS }
%     \and
%     \inferrule*[right=\scriptsize h-set]
%     { q:\SS \vdashtheta{v} q : \SS \\ Q \in \Theta }
%     { \Gamma, q:\SS \vdashtheta{c} \forall v.\{ v = \emptyset \}\, \mathtt{set}\,Q\,v \,\{\lambda \_. Q = \emptyset\}: \UU }
% }
% { \Gamma \vdashtheta{c} \{ \top \}\, \letin{q=\mathtt{initQ}}{\mathtt{set}\,Q\,q} \,\{\lambda \_.\mathtt{coherent}(Q, rand)\} : \UU }
% \and
% \inferrule*[right=\scriptsize h-lam]
% {
%     \Gamma, u: \UU \vdashtheta{c} \{ \top \}\,\letin{q=\mathtt{initQ}}{\mathtt{set}\,Q\,q} \,\{\lambda\_.\mathtt{coherent}(Q, rand)\} : \UU
% }
% {
%     \Gamma \vdashtheta{v} \lambda u.\letin{q=\mathtt{initQ}}{\mathtt{set}\,Q\,q} : \UU \underset{\lambda u.\top, \lambda u.\lambda \_.\mathtt{coherent}(Q, rand)}{\multimap} \UU
% }

% \inferrule*[right=\scriptsize h-cons, left=\scriptsize $\mathcal{D}_{initQ}$]
% {
%     \inferrule*[right=\scriptsize h-eta]
%     { \vdashtheta{v} \mathtt{initQ} : \SS }
%     { \vdashtheta{c} \{\top\} \, \return \mathtt{initQ} \, \{\lambda v. v = \mathtt{initQ}\} : \SS }
%     \quad
%     \forall v. v = \mathtt{initQ} \vdash \mathtt{coherent}(v, rand) \quad \text{(Lemma~\ref{lemma:coherent-empty-q})}
% }
% {
%     \vdashtheta{c} \{\top\}\, \return \mathtt{initQ} \,\{\lambda v.\mathtt{coherent}(v, rand)\} : \SS
% }
% \and
% \inferrule* [right=\scriptsize h-set, left=\scriptsize $\mathcal{D}_{setQ}$]
% {
%     q : \SS \vdashtheta{v} q : \SS \\ Q \in \Theta
% }
% {
%     q: \SS \vdashtheta{c} \{\mathtt{coherent}(q, rand)\}\, \mathtt{set}\,Q\,q \,\{\lambda\_.\mathtt{coherent}(Q, rand)\} : \UU
% }
% \and
% \inferrule* [right=\scriptsize h-let, left=\scriptsize $\mathcal{D}_{Qledger}$]
% {
%     \mathcal{D}_{initQ} \\ \mathcal{D}_{setQ}
% }
% {
%     \vdashtheta{c} \{\top\}\, \lambda u.\letin{q=\mathtt{initQ}}{\mathtt{set}\,Q\,q} \,\{\lambda\_.\mathtt{coherent}(Q, rand)\} : \UU
% }
% \and
% \inferrule*[right=\scriptsize h-cons, left=\scriptsize $\mathcal{D}_{initR}$]
% {
%     \inferrule*[right=\scriptsize h-eta]
%     { \vdashtheta{v} \mathtt{initTable} : \SS }
%     { \vdashtheta{c} \{\top\} \, \return \mathtt{initTable} \, \{\lambda v. v = \mathtt{initTable}\} : \SS }
%     \quad
%     \forall v. v = \mathtt{initTable} \vdash \mathtt{coherent}(Q, v) \quad \text{(Lemma~\ref{lemma:coherent-empty-q})}
% }
% {
%     \vdashtheta{c} \{\mathtt{coherent}(Q, rand)\}\, \return \mathtt{initTable} \,\{\lambda v.\mathtt{coherent}(Q, v)\} : \SS
% }
% \and
% \inferrule* [right=\scriptsize h-set, left=\scriptsize $\mathcal{D}_{setR}$]
% {
%     tab : \SS \vdashtheta{v} tab : \SS \\ rand \in \Theta
% }
% {
%     tab: \SS \vdashtheta{c} \{\mathtt{coherent}(Q, tab)\}\, \mathtt{set}\,rand\,tab \,\{\lambda\_.\mathtt{coherent}(Q, rand)\} : \UU
% }
% \and
% \inferrule* [right=\scriptsize h-let, left=\scriptsize $\mathcal{D}_{Rledger}$]
% {
%     \mathcal{D}_{initR} \\ \mathcal{D}_{setR}
% }
% {
%     \vdashtheta{c} \{\mathtt{coherent}(Q, rand)\}\, \letin{tab=\mathtt{initTable}}{\mathtt{set}\,rand\,tab} \,\{\lambda\_.\mathtt{coherent}(Q, rand)\} : \UU
% }
% \and
% \inferrule*[right=\scriptsize h-eta, left=\scriptsize $\mathcal{D}_{retZero}$]
% { \vdashtheta{v} \mathtt{zero} : \SS }
% { \vdashtheta{c} \{ \mathtt{coherent}(Q, rand) \} \, \return \mathtt{zero} \, \{ \lambda \_.\mathtt{coherent}(Q, rand) \} : \SS }
% \and
% \inferrule* [right=\scriptsize h-let, left=\scriptsize $\mathcal{D}_{Gen}$]
% {
%     \mathcal{D}_{Qledger} \\
%     \inferrule* [right=\scriptsize h-let]
%     {
%         \mathcal{D}_{Rledger} \\
%         \mathcal{D}_{retZero}
%     }
%     {
%         \vdashtheta{c} \{\mathtt{coherent}(Q, rand)\}\, \\\\
%         \letin{tab=\mathtt{initTable}}{\letin{\_=\mathtt{set}\,rand\,tab}{\return \mathtt{zero}}} \\\\ 
%         \,\{\lambda \_.\mathtt{coherent}(Q, rand)\} : \SS
%     }
% }
% {
%     \vdashtheta{c} \{\top\}\, 
%     \widetilde{Gen} \star
%     \,\{\lambda \_.\mathtt{coherent}(Q, rand)\} : \SS
% }
% \end{mathpar}
% \end{center}

Finally we state and prove Theorem~\ref{theorem:tilde-macforge-sim-false} that completes the proof.

\begin{theorem}\label{theorem:tilde-macforge-sim-false}
\begin{align*}
\widetilde{MacForge}\simcontext \false
\end{align*}
\end{theorem}

\begin{proof}

\small{
\begin{align*}
    \widetilde{MacForge} &\quad\triangleq\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\letin{o = \widetilde{Oracle}\, k}{} \\
        &\letin{\lis{m,tag} = Adv\, o}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag)}
    \end{aligned}
    \\[2ex]
    &\quad\triangleq\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\mathtt{let}\, o = \left(
        \begin{aligned}
            &\return(\lambda k.\return \bang(\return \lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m}))
        \end{aligned}
        \right) k \,\mathtt{in}\\
        &\letin{\lis{m,tag} = Adv\, o}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag)}
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\mathtt{let}\, o = \left(
        \begin{aligned}
            &\return \bang(\return \lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m})
        \end{aligned}
        \right)\mathtt{in}\\
        &\letin{\lis{m,tag} = Adv\, o}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\letin{x=\mathtt{der}(o)\, m}{\mathtt{eq}(x,tag)}
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc4}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\mathtt{let}\, \lis{m,tag} = Adv \left(
        \begin{aligned}
            &\bang(\return \lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m})
        \end{aligned}
        \right)\mathtt{in} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\,c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\mathtt{let}\, x=\mathtt{der} \left(
        \begin{aligned}
            &\bang(\return \lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m})
        \end{aligned}
        \right) m \,\mathtt{in} \\
        &\mathtt{eq}(x,tag)
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc3}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\mathtt{let}\, \lis{m,tag} = Adv \left(
        \begin{aligned}
            &\bang(\return \lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m})
        \end{aligned}
        \right)\,\mathtt{in} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\mathtt{let}\, x=\left(
        \begin{aligned}
            &\return(\lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m})
        \end{aligned}
        \right) m \,\mathtt{in} \\
        &\mathtt{eq}(x,tag)
    \end{aligned}
    \\[2ex]
    &\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\mathtt{let}\, \lis{m,tag} = Adv \left(
        \begin{aligned}
            &\bang(\return \lambda m. \\
            &\quad \letin{q=\get Q}{} \\
            &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\quad \mathtt{else}\ \mathtt{skip}; \\
            &\quad \mathit{\widetilde{Mac}}\lis{k,m})
        \end{aligned}
        \right)\mathtt{in} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
        &\mathtt{else}\,\mathtt{let}\, x=\left(
        \begin{aligned}
            &\letin{q=\get Q}{} \\
            &\letin{t=\mathtt{contains}(q,m)}{} \\
            &\mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \,\mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
            &\mathtt{else}\ \mathtt{skip}; \\
            &\mathit{\widetilde{Mac}}\lis{k,m}
        \end{aligned}
        \right)\mathtt{in} \\
        &\mathtt{eq}(x,tag)
    \end{aligned}
\end{align*}
}

\normalsize

At this stage, the adversary is given access to a term that can be evaluated for at most a polynomial number of queries~$q$, and it must output a candidate forgery. In the else branch, the oracle is reduced to an expression that applies a sequence of operations to the query ledger and returns the output of $\widetilde{Mac}$. We focus on the lower part of the term, namely the else branch. This branch is unrolled by expanding the definition of $\widetilde{Mac}$. By repeated applications of Equation~\ref{letAss}, the resulting term is rearranged. Applying Equation~\ref{iflet} pushes the binding of the variable $x$ inside the conditional. The resulting term simplifies reasoning in terms of Hoare assertions.

\scriptsize{
\begin{align*}
&\quad\triangleq\quad
\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\mathtt{let}\, \lis{m,tag} = Adv \left(
    \begin{aligned}
        &\bang(\return \lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad(\lambda e. \\
        &\quad\letin{\lis{k,m} = e}{} \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad \return x \\
        &\quad\mathtt{else} \\
        &\quad\quad \letin{x = \random}{} \\
        &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad \return x) \lis{k,m})
    \end{aligned}
    \right)\mathtt{in} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
    &\mathtt{else}\,\mathtt{let}\, x=\left(
    \begin{aligned}
        &\letin{q=\get Q}{} \\
        &\letin{t=\mathtt{contains}(q,m)}{} \\
        &\mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\mathtt{else}\,\mathtt{skip}; \\
        &(\lambda e. \\
        &\letin{\lis{k,m} = e}{} \\
        &\letin{tab = \mathtt{get}\,rand}{} \\
        &\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\mathtt{if}\, b\, \mathtt{then} \\
        &\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad \return x \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad \mathtt{set}\,rand\,newtab; \\
        &\quad \return x) \lis{k,m}  
    \end{aligned}
    \right)\mathtt{in} \\
    &\mathtt{eq}(x,tag)
\end{aligned}
\\[2ex]
\quad\overset{\text{~\ref{reduc2}}}{\simcontext}\quad&
\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\mathtt{let}\, \lis{m,tag} = Adv \left(
    \begin{aligned}
        &\bang(\return \lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad\letin{\lis{k,m} = \lis{k,m}}{} \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad \return x \\
        &\quad\mathtt{else} \\
        &\quad\quad \letin{x = \random}{} \\
        &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad \return x) 
    \end{aligned}
    \right)\mathtt{in} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
    &\mathtt{else}\,\mathtt{let}\, x=\left(
    \begin{aligned}
        &\letin{q=\get Q}{} \\
        &\letin{t=\mathtt{contains}(q,m)}{} \\
        &\mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
        &\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\mathtt{else}\,\mathtt{skip}; \\
        &\letin{\lis{k,m} = \lis{k,m}}{} \\
        &\letin{tab = \mathtt{get}\,rand}{} \\
        &\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\mathtt{if}\, b\, \mathtt{then} \\
        &\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad \return x \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad \mathtt{set}\,rand\,newtab; \\
        &\quad \return x
    \end{aligned}
    \right)\mathtt{in} \\
    &\mathtt{eq}(x,tag)
\end{aligned}
\\[2ex]
\quad\overset{\text{~\ref{reduc5}}}{\simcontext}\quad&
\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\mathtt{let}\, \lis{m,tag} = Adv \left(
    \begin{aligned}
        &\bang(\return \lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad \return x \\
        &\quad\mathtt{else} \\
        &\quad\quad \letin{x = \random}{} \\
        &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad \return x)
    \end{aligned}
    \right)\mathtt{in} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
    &\mathtt{else}\,\mathtt{let}\, x=\left(
    \begin{aligned}
        &\letin{q=\get Q}{} \\
        &\letin{t=\mathtt{contains}(q,m)}{} \\
        &\mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
        &\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\mathtt{else}\,\mathtt{skip}; \\
        &\letin{tab = \mathtt{get}\,rand}{} \\
        &\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\mathtt{if}\, b\, \mathtt{then} \\
        &\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad \return x \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad \mathtt{set}\,rand\,newtab; \\
        &\quad \return x
    \end{aligned}
    \right)\mathtt{in} \\
    &\mathtt{eq}(x,tag)
\end{aligned}
\end{align*}
}

\normalsize

At this point in the equational reasoning, the term passed as argument to $Adv$ cannot be reduced further. From now on, we denote the argument of $Adv$ as $O$ (for oracle) in order to keep the full term more compact.

\small{
\begin{align}
    O \quad\triangleq\quad \left( 
    \begin{aligned}
        &\bang(\return \lambda m. \\
        &\quad \letin{q=\get Q}{} \\
        &\quad \letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \, \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
        &\quad \mathtt{else}\ \mathtt{skip}; \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad\mathtt{if}\, b\, \mathtt{then} \\
        &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad\quad \return x \\
        &\quad\mathtt{else} \\
        &\quad\quad \letin{x = \random}{} \\
        &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad\quad \mathtt{set}\,rand\,newtab; \\
        &\quad\quad \return x)
    \end{aligned}
    \right) \label{eq:oracle-compacted}  
\end{align}
}

\footnotesize{
\begin{align*}
\quad\overset{\text{~\ref{letAss}}}{\simcontext}\quad&
\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\letin{\lis{m,tag} = Adv\,O}{} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
    &\mathtt{else}\\
    &\quad\letin{q=\get Q}{} \\
    &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
    &\quad\mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
    &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
    &\quad\mathtt{else}\,\mathtt{skip}; \\
    &\quad\letin{tab = \mathtt{get}\,rand}{} \\
    &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    &\quad\mathtt{let}\, x=\left(
    \begin{aligned}
        &\mathtt{if}\, b\, \mathtt{then} \\
        &\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
        &\quad \return x \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad \mathtt{set}\,rand\,newtab; \\
        &\quad \return x
    \end{aligned}
    \right)\mathtt{in} \\
    &\quad\mathtt{eq}(x,tag)
\end{aligned}
\\[2ex]
\quad\overset{\text{~\ref{iflet}}}{\simcontext}\quad&
\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\letin{\lis{m,tag} = Adv\,O}{} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    &\mathtt{if}\, c\, \mathtt{then}\, \return \false \\
    &\mathtt{else}\\
    &\quad\letin{q=\get Q}{} \\
    &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
    &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
    &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q,m) \\
    &\quad\mathtt{else}\,\mathtt{skip}; \\
    &\quad\letin{tab = \mathtt{get}\,rand}{} \\
    &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    &\quad\mathtt{if}\, b\, \mathtt{then} \\
    &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
    &\quad\quad \mathtt{eq}(x,tag) \\
    &\quad\mathtt{else} \\
    &\quad\quad \letin{x = \random}{} \\
    &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
    &\quad\quad \mathtt{set}\,rand\,newtab; \\
    &\quad\quad \mathtt{eq}(x,tag)
\end{aligned}
\end{align*}
}

\normalsize

Now we need to refine the typing of the term $Adv$ with respect of the typing rules introduced in Figure~\ref{fig:hl-typing-rules}. The original type designated for $Adv$ was $\vdash_c^\varnothing Adv: \bang_{q}\bigl(\SS[i] \multimap \SS[i]\bigr) \multimap (\SS[i] \otimes \SS[i])$. We lift the type of $Adv$ in the Hoare typing system using the map $\mu$ defined in Equation~\ref{eq:mu}. By fixing a proposition $P$ as $\mathtt{coherent}(Q, rand)$ and by applying Lemma~\ref{lemma:hoare-lifting}, the type of $Adv$ is lifted as follows:
\begin{align*}
&\Gamma \vdash_c^\varnothing Adv: \bang_{q}\bigl(\SS[i] \multimap \SS[i]\bigr) \multimap (\SS[i] \otimes \SS[i])
\end{align*}
and
\begin{align*}
&\mu(\Gamma, P) \vdash^\Xi_{c} Adv: \mu(\bang_{q}\bigl(\SS[i] \multimap \SS[i]\bigr) \multimap (\SS[i] \otimes \SS[i])).
\end{align*}
Which expands as
\begin{align*}
&\mu\bigl(\bang_{q}(\SS[i] \multimap \SS[i]) \multimap (\SS[i] \otimes \SS[i]), P\bigr) \\
&= \mu\bigl(\bang_{q}(\SS[i] \multimap \SS[i]), P\bigr) \underset{P, \lambda \_.P}{\multimap} \mu(\SS[i] \otimes \SS[i], P) \\
&= \bang_{q}^{P,P} \mu(\SS[i] \multimap \SS[i], P) \underset{P, \lambda \_.P}{\multimap} (\SS[i] \otimes \SS[i]) \\
&= \bang_{q}^{P,P} \Bigl(\SS[i] \underset{P, \lambda \_.P}{\multimap} \SS[i]\Bigr) \underset{P, \lambda \_.P}{\multimap} (\SS[i] \otimes \SS[i]).
\end{align*}
The application of Lemma~\ref{lemma:hoare-lifting} is justified by the type of $Adv$ and by the fact that the typing derivation is performed in the empty context, so that $\mathcal{R}(P) \cap \varnothing = \emptyset$. Moreover, $\Theta = \varnothing$ and $\Xi = \varnothing$. Under these conditions, the type of $Adv$ guarantees that $P$ is preserved as an invariant. By applying the lifted typing judgment for $Adv$ and combining it with Lemma~\ref{lemma:oracle-coherence}, we can propagate the assertion $\mathtt{coherent}(Q, rand)$ down into the term. Note that the term $O$, defined in Equation~\ref{eq:oracle-compacted} is exactly the same as term $M$ in Lemma~\ref{lemma:oracle-coherence}.

\scriptsize{
\begin{align*}
&\begin{aligned}
    &\color{blue} \{ \top \} \\
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\color{blue} \{ \mathtt{coherent}(Q, rand) \} \quad \text{Lemma~\ref{lemma:gen-coherence}} \\
    &\letin{\lis{m,tag} = Adv\,O}{} \\
    &\color{blue} \{ \mathtt{coherent}(Q, rand) \} \quad \text{Lemmatas~\ref{lemma:hoare-lifting}, \ref{lemma:oracle-coherence}} \\
    &\letin{queries = \get Q}{} \\
    &\color{blue} \{ \mathtt{coherent}(queries, rand) \land (queries=Q)\} \quad \text{\ref{rule:h-get}} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    &\color{blue} \{ \mathtt{coherent}(queries, rand) \land (queries=Q) \land c=(m\in queries) \} \quad \text{\ref{rule:h-fun}, \ref{rule:h-cons}} \\
    & \mathtt{if}\, c\, \mathtt{then} \\
    &\quad \color{blue} \{ \mathtt{coherent}(queries, rand) \land (queries=Q) \land (m\in queries) \} \quad \text{\ref{rule:h-case}, \ref{rule:h-cons}} \\
    &\quad \return \false \\
    &\quad \color{blue} \{ \lambda b.\mathtt{coherent}(Q, rand) \land (m\in Q) \} \quad \text{\ref{rule:h-cons}, \ref{rule:h-eta}} \\
    &\mathtt{else} \\
    &\quad \color{blue} \{ \mathtt{coherent}(queries, rand) \land (queries=Q) \land (m \not\in queries) \} \quad \text{\ref{rule:h-case}, \ref{rule:h-cons}} \\
    &\quad\letin{q=\get Q}{} \\
    &\quad \color{blue} \{ \mathtt{coherent}(queries, rand) \land (queries=Q) \land (m \not\in queries) \land (q = Q) \} \quad \text{\ref{rule:h-get}, \ref{rule:h-cons}} \\
    &\qquad \implies \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \not\in q) \} \quad \text{\ref{rule:h-cons}} \\
    &\quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \not\in q) \} \quad \text{\ref{rule:h-cons}} \\
    &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
    &\quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \not\in q) \land t = (m \in q)\} \quad \text{\ref{rule:h-fun}} \\
    &\qquad \implies \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \not\in q) \land t = \false \} \quad \text{\ref{rule:h-cons}} \\
    &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
    &\qquad \color{blue} \{ \mathtt{coherent}(q, rand) \land (q=Q) \land (m \not\in q) \} \quad \text{\ref{rule:h-case}, \ref{rule:h-cons}} \\
    &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m) \\
    &\qquad \color{blue} \{ \mathtt{coherent}(q, rand) \land (Q = q \cup \{m\}) \land (m \notin q) \} \quad \text{\ref{rule:h-fun}, \ref{rule:h-set}, \ref{rule:h-cons} } \\
    &\quad \mathtt{else} \\ 
    &\qquad \mathtt{skip}; \\
    &\quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (Q = q \cup \{m\}) \land (m \notin q) \} \quad \text{} \\
    &\quad\letin{tab = \mathtt{get}\,rand}{} \\
    &\quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (Q = q \cup \{m\}) \land (m \notin q) \land (tab=rand) \} \quad \text{\ref{rule:h-get}}\\
    &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    &\quad \color{blue} \{ \mathtt{coherent}(q, rand) \land (Q = q \cup \{m\}) \land (m \notin q) \land (tab=rand) \land (b=(m \in tab)) \} \quad \text{\ref{rule:h-fun}} \\
    &\qquad \implies \color{blue} \{ b = \false \} \quad \text{\ref{rule:h-cons}} \\
    &\quad\mathtt{if}\, b\, \mathtt{then} \\
    &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
    &\quad\quad \mathtt{eq}(x,tag) \\
    &\quad\mathtt{else} \\
    &\quad\quad \letin{x = \random}{} \\
    &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
    &\quad\quad \mathtt{set}\,rand\,newtab; \\
    &\quad\quad \mathtt{eq}(x,tag)
\end{aligned}
\end{align*}
}

\normalsize

At this point we do not introduce assertions any further. We observe instead that two assertion entails that the guard of a conditional is invariably evaluated to true or false. In particular, one of the two conditionals determines whether the oracle outputs a uniformly random bitstring upon receiving a fresh message generated by $Adv$, or whether it returns a predetermined bitstring stored in the ledger $rand$. 

The message produced by the adversary must be fresh, hence it cannot appear in the ledger $rand$, which records all outputs of the random oracle. Moreover, it cannot appear in the set $Q$ at this stage either, because the adversary's objective is precisely to produce a fresh message–tag pair. Since the oracle first updates the query ledger and only then invokes $\widetilde{Mac}$ to recompute the tag and verify that the tag provided by the adversary is valid, by tracking the initial query ledger, under the precondition that it is consistent with the $rand$ ledger and that it does not contain the message $m$, we can conclude that $m$ does not appear in the $rand$ ledger as well. Consequently, the conditional collapses to the branch that returns a random bitstring. We apply Equation~\ref{ifFalse} and Equation~\ref{ifTrue} to further rewrite the term.

\footnotesize{
\begin{align*}
&\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\letin{\lis{m,tag} = Adv\,O}{} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    & \mathtt{if}\, c\, \mathtt{then} \\
    &\quad \return \false \\
    &\mathtt{else} \\
    &\quad\letin{q=\get Q}{} \\
    &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
    &\quad \color{blue} \{ t = \false \} \\
    &\quad \mathtt{if}\ \mathtt{not}(t)\,\mathtt{then} \\
    &\quad\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m) \\
    &\quad \mathtt{else} \\ 
    &\qquad \mathtt{skip}; \\
    &\quad\letin{tab = \mathtt{get}\,rand}{} \\
    &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    &\quad  \color{blue} \{ b = \false \} \\
    &\quad\mathtt{if}\, b\, \mathtt{then} \\
    &\quad\quad \letin{x = \mathtt{getValue}(tab, m)}{} \\
    &\quad\quad \mathtt{eq}(x,tag) \\
    &\quad\mathtt{else} \\
    &\quad\quad \letin{x = \random}{} \\
    &\quad\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
    &\quad\quad \mathtt{set}\,rand\,newtab; \\
    &\quad\quad \mathtt{eq}(x,tag)
\end{aligned}
\\[2ex]
&\quad\overset{\text{~\ref{ifFalse}~\ref{ifTrue}}}{\sim}\quad
\begin{aligned}
    &\letin{k = \widetilde{Gen} \star}{} \\
    &\letin{\lis{m,tag} = Adv\,O}{} \\
    &\letin{queries = \get Q}{} \\
    &\letin{c=\mathtt{contains}(queries, m)}{} \\
    & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
    &\mathtt{else} \\
    &\quad\letin{q=\get Q}{} \\
    &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
    &\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m); \\
    &\quad\letin{tab = \mathtt{get}\,rand}{} \\
    &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
    &\quad \letin{x = \random}{} \\
    &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
    &\quad \mathtt{set}\,rand\,newtab; \\
    &\quad \mathtt{eq}(x,tag)
\end{aligned}
\end{align*}
}

\normalsize

In order to conclude, we leave assertions aside and we start to simplify the term going back to the notion of $\Theta$-contextual indistinguishability of \cite{lago2024computationalindistinguishabilitylogicalrelations}. Recall that $O$ is not a free variable but it is the term defined in Equation~\ref{eq:oracle-compacted}.

\small{
\begin{align*}
    &\begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
        &\mathtt{else} \\
        &\quad\letin{q=\get Q}{} \\
        &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m); \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad \letin{x = \random}{} \\
        &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad \mathtt{set}\,rand\,newtab; \\
        &\quad \mathtt{eq}(x,tag)
    \end{aligned}
    % \\[2ex]
    % &\quad\underset{\text{\ref{letCom}}}{\simcontext}\quad
    \quad\underset{\text{\ref{letCom}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad\letin{q=\get Q}{} \\
        &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
        &\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m); \\
        &\quad\letin{tab = \mathtt{get}\,rand}{} \\
        &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
        &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
        &\quad \mathtt{set}\,rand\,newtab; \\
        &\quad \mathtt{eq}(x,tag)
    \end{aligned}
    \\[2ex]
    &\quad\underset{\text{\ref{letAss}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \mathtt{let}\,{\_=} \left(
        \begin{aligned}
            &\quad\letin{q=\get Q}{} \\
            &\quad\letin{t=\mathtt{contains}(q,m)}{} \\
            &\quad \mathtt{set}\,Q\;\mathtt{modifyQ}(q, m); \\
            &\quad\letin{tab = \mathtt{get}\,rand}{} \\
            &\quad\letin{b = \mathtt{isdefined}(tab, m)}{} \\
            &\quad \letin{newtab = \mathtt{modify}(tab, m, x)}{} \\
            &\quad \mathtt{set}\,rand\,newtab
        \end{aligned}
        \right) \mathtt{in} \\
        &\quad \mathtt{eq}(x,tag)
    \end{aligned}
    \\[2ex]
    &\quad\underset{\text{\ref{safeDisc1}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{k = \widetilde{Gen} \star}{} \\
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \mathtt{eq}(x,tag)
    \end{aligned}
    \\[2ex]
    &\quad\underset{\text{\ref{safeDisc1}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
        &\mathtt{else} \\
        &\quad \letin{x = \random}{} \\
        &\quad \mathtt{eq}(x,tag)
    \end{aligned}
    \quad\underset{\text{\ref{randF}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \mathtt{if}\, c\, \mathtt{then} \, \return \false \\
        &\mathtt{else} \, \return \false
    \end{aligned}
    \\[2ex]
    &\quad\underset{\text{\ref{ifConst}}}{\simcontext}\quad
    \begin{aligned}
        &\letin{\lis{m,tag} = Adv\,O}{} \\
        &\letin{queries = \get Q}{} \\
        &\letin{c=\mathtt{contains}(queries, m)}{} \\
        & \return \false
    \end{aligned}
    \quad\underset{\text{\ref{safeDisc1}}}{\simcontext}\quad
    \begin{aligned}
        &\return \false
    \end{aligned}
\end{align*}
}
\end{proof}

Summing up, we return to the diagram in Figure~\ref{fig:proofscheme}, which provides a structured view of the overall security proof in the $\lamBLL$ framework. We first establish the key equivalences involving the distinguishers. In particular, Theorem~\ref{theorem:tildeD-sim-tildeMACForge} shows that $\widetilde{D} \sim \widetilde{MacForge}$, while Theorem~\ref{theorem:DF-sim-MacForgeF} establishes that $D^F \simcontext \mathit{MacForge}^F$. The final step of the proof is to show that $\widetilde{MacForge} \simcontext \false$, corresponding to the top-right part of the diagram. To obtain this result, we first prove that certain invariants hold throughout the subterms of the experiment. This is done using the Hoare logic introduced in Section~\ref{sec:extensions-lambda-bll}, which extends the original typing system of $\lamBLL$. Decorating the terms with assertions allows us to establish properties about the state of the experiment. These properties enable the application of the equations introduced in Section~\ref{sec:extensions-lambda-bll}, which collapse unreachable conditional branches. Using Lemma~\ref{lemma:hoare-lifting}, we then reason about the typing of the adversary. Once we show in Lemma~\ref{lemma:oracle-coherence} that the oracle with which the adversary interacts satisfies a given invariant, the type of the adversary ensures that the invariant is preserved. Using the typing rules for assertions, we propagate the invariant through the term and apply the equations mentioned above. Finally, we return to the original contextual indistinguishability reasoning. By mapping the behavior of each component in the proof to its counterpart through the logical relation, we conclude that the scheme satisfies the required indistinguishability guarantees.

