\rhead[\fancyplain{}{\bfseries CONCLUSIONS}]{\fancyplain{}{\bfseries\thepage}}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries CONCLUSIONS}}
\chapter*{Conclusions}
\addcontentsline{toc}{chapter}{Conclusions}

This thesis explored the use of programming-language techniques for the formalization of cryptographic proofs in the computational model. In particular, the work focused on the probabilistic lambda calculus $\lamBLL$, whose graded type system characterizes probabilistic polynomial-time computation. The objective was to assess whether such a framework could support the formalization of reduction-based security proofs commonly used in computational cryptography. To this end, we introduced an extension of the $\lamBLL$ type system inspired by Hoare logic, allowing terms to be annotated with assertions describing properties of program states before and after execution. This enriched typing system makes it possible to reason about invariants and stateful components that naturally arise in cryptographic constructions, such as the tables used to simulate function oracles or the set of queries performed by an adversary to an oracle. Within this framework, we formalized the security proof of message authentication codes constructed from pseudorandom functions. In particular, we modeled the MacForge experiment and proved existential unforgeability under chosen-message attacks by reproducing the classical reduction to pseudorandom functions. The proof was expressed equationally inside the calculus, relying on the extended set of functional symbols and equations together with the Hoare-style reasoning principles. This work demonstrates that $\lamBLL$ provides a suitable foundation for representing cryptographic experiments and reduction-based arguments while statically enforcing polynomial-time resource bounds. The introduction of Hoare-style assertions further enables structured reasoning about stateful computations and invariants during execution.

Despite the fact that we obtained a formal proof, several challenges remain. The current framework requires a significant amount of manual reasoning and explicit modeling of cryptographic components. Improving automation and developing higher-level abstractions would facilitate the use of the system for larger and more complex cryptographic proofs. 

Overall, this work provides evidence that programming-language techniques and in particular type systems for resource-bounded computation, can offer a promising foundation for the formal verification of proofs in computational cryptography.