\renewcommand{\chaptermark}[1]{\markright{\thechapter \ #1}{}}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries\rightmark}}
\appendix
\chapter{Useful Function Symbols and Equations in lambda-BLL}\label{app:1}

Table~\ref{table:function-symbols} introduces a collection of function symbols operating over boolean values and bitstrings, specifically tailored to facilitate random generation, basic data manipulation, and operations on a specialized ledger structure. While individually simple, these symbols collectively provide a robust framework for modeling both pure computation and stateful interaction within $\lamBLL$. The symbols in Table~\ref{table:function-symbols} are divided in three groups based on their functional roles. It's important to note that $\lamBLL$, each function symbol in $\mathcal{F}$ denotes a mapping into a probability distribution over values of its respective output type, as established in Section 1 of \cite{lago2024computationalindistinguishabilitylogicalrelations}. Consequently, when a function symbol is interpreted as returning a specific value $v$, it corresponds to the Dirac distribution centered at $v$ (denoted $\delta_v$), which assigns probability $1$ to $v$ and $0$ to all other outcomes.

\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.0}
\begin{footnotesize}
\begin{xltabular}{\textwidth}{|@{\hskip 1.5pt}c|l|@{\hskip 1.5pt}X@{\hskip 1.5pt}|}
    \hline
    \textbf{Function Symbol} & \multicolumn{1}{c|}{\textbf{Type}} & \multicolumn{1}{c|}{\textbf{Interpretation}} \\ \hline
    \endfirsthead
    \hline
    \textbf{Function Symbol} & \multicolumn{1}{c|}{\textbf{Type}} & \multicolumn{1}{c|}{\textbf{Interpretation}} \\ \hline
    \endhead
    \multicolumn{3}{|c|}{\textbf{Function Symbols Generating Random Values}} \\ \hline
    $\random$ & $\SS[i]$ & Function generating a random bitstring in $\{0,1\}^i$. \\ \hline
    
    \multicolumn{3}{|c|}{\textbf{Function Symbols for Manipulating Bitstrings and Boolean Values}} \\ \hline
    $\mathtt{not}$ & $\BB \to \BB$ & Function computing the negation of a boolean value. \\ \hline
    $\mathtt{xor}$ & $\SS[i] \to \SS[i]$ & Function computing the bitwise exclusive-or of two binary strings with the same length. \\ \hline
    $\mathtt{zero}$ & $\SS[i]$ & Function that returns a bitstring consisting solely of zeros. \\ \hline
    
    \multicolumn{3}{|c|}{\textbf{Function Symbols for Ledger Manipulation}} \\ \hline
    $\mathtt{initTable}_p$ & $\SS[p\times(2i+1)]$ & Function that initializes a ledger by generating a string of length $p\times(2i+1)$ filled entirely with zeros. \\ \hline
    
    $\mathtt{isdefined_p}$ & $\SS[p\times(2i+ 1)]\times\SS[i]\to\BB$ & Function that takes as input a ledger and a string of length $i$, and returns true if the row, whose first $i$ bits match the input string, is active and false otherwise. \\ \hline
    
    $\mathtt{getValue_p}$ & $\SS[p\times(2i+ 1)]\times\SS[i]\to\SS[i]$ & Function that, given a ledger and an input string of length $i$, searches for a row in the ledger whose first $i$ bits are equal to the input string and whose final bit is set to 1 (indicating that the row is active), and returns the substring spanning from position $i+1$ to position $2i$ within the selected row. If no matching active row is found, it returns a dummy value. \\ \hline
    
    $\mathtt{modify}_p$ & $\SS[p\times(2i+1)]$ & Function that, given a ledger and two input strings of length $i$, identifies the first inactive row in the ledger, inserts into that row the concatenation of the two strings followed by a bit set to 1 (indicating activation), and returns the resulting updated ledger. \\ \hline
    \noalign{\smallskip}
    \noalign{\smallskip}
    \noalign{\smallskip}
    \noalign{\smallskip}
    \noalign{\smallskip}
    \caption{Table of function symbols and their interpretations as presented in \cite{lago2024computationalindistinguishabilitylogicalrelations}.}
    \label{table:function-symbols}
\end{xltabular}
\end{footnotesize}

The equations presented in Figure~\ref{fig:equations} establish a set of equivalences between $\lamBLL$ terms, which are sound with respect to the notion of $\Theta$-contextual indistinguishability introduced in \cite{lago2024computationalindistinguishabilitylogicalrelations} and denoted here by $\sim_{\Theta\subseteq\Xi}$. These equations characterize behavioral equivalence for terms that remain indistinguishable to any probabilistic polynomial-time context, restricted by a specific set of observable references.

\begin{figure}[htpb]
    \centering
    \footnotesize
    \begin{align}
    \letin{x=M}{\return x} &\sim_{\Theta\subseteq\Xi}\return M\label{reduc1}\tag{$\mathtt{reduc1}$} \\[2ex]
    (\return \lambda x.M)Z &\sim_{\Theta\subseteq\Xi}M[Z/x]\label{reduc2}\tag{$\mathtt{reduc2}$} \\[2ex]
    \mathtt{der}(\bang M)&\sim_{\Theta\subseteq\Xi}M\label{reduc3}\tag{$\mathtt{reduc3}$} \\[2ex]
    \letin{x=\return V}{M}&\sim_{\Theta\subseteq\Xi}M[V/x] \label{reduc4}\tag{$\mathtt{reduc4}$} \\[2ex]
    \letin{\lis{x,y}=\lis{Z,Z'}}{M}&\sim_{\Theta\subseteq\Xi}M[Z/x,Z'/y]\label{reduc5}\tag{$\mathtt{reduc5}$} \\[2ex]
    \left(
    \begin{aligned}
        &\letin{r=\random}{} \\
        &\letin{z=\mathtt{isdefined}(y,r)}{} \\
        &\ifthen{z}{M}{N}
    \end{aligned}\right) &\sim_{\Theta\subseteq\Xi} \letin{r=\random}{N}\label{randT}\tag{$\mathtt{randT}$}
    \\[2ex]
    \letin{x=M_1}{\letin{y=M_2}{N}}{} &\sim_{\Theta\subseteq\Xi} \letin{y=(\letin{x=M_1}{M_2)}}{N} \label{letAss}\tag{$\mathtt{letAss}$} \\
    \text{if}\ x &\not\in FV(N) \nonumber
    \\[2ex]
    \letin{x_1=M_1}{\letin{x_2=M_2}{N}} &\sim_{\Theta\subseteq\Xi} \letin{x_2=M_2}{\letin{x_1=M_1}{N}}\tag{$\mathtt{letCom}$}\label{letCom}
    \\[2ex]
    \letin{x=M}{N}&\sim_{\Theta\subseteq\Xi}N \tag{$\mathtt{safeDisc1}$}\label{safeDisc1} \\
    \text{if}\ x \not\in FV(N) \ \text{and}\ \mathcal{R}(M) &\cap \mathcal{R}(N) = \emptyset \ \text{and}\ \nonumber \\ \mathcal{R}(M) \cap \Theta = \emptyset \ \text{and} &HOFV(M) = \emptyset \nonumber
    \\[2ex]
    \letin{\lis{x,y}=M}{N}&\sim_{\Theta\subseteq\Xi}N \tag{$\mathtt{safeDisc2}$}\label{safeDisc2} \\
    \text{if}\ x,y \not\in FV(N) \ \text{and}\ \mathcal{R}(M) &\cap \mathcal{R}(N) = \emptyset \ \text{and}\ \nonumber \\ \mathcal{R}(M) \cap \Theta = \emptyset \ \text{and} &HOFV(M) = \emptyset \nonumber 
    \end{align}
    \caption{Equations in $\lamBLL$ as presented in \cite{lago2024computationalindistinguishabilitylogicalrelations}. Let $\mathcal{R}(M)$ be the set of references accessed by the term $M$. The equations in this figure hold for all reference contexts $\Theta, \Xi$ such that $\Theta \subseteq \Xi$, except where stated otherwise.}
    \label{fig:equations}
\end{figure}